---
title: "Bayesian Modeling of STIR Effects on Water Quality"
author: "AJ Brown"
output:
  html_document:
    toc: true
    toc_depth: 3
    number_sections: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = TRUE,
  message = TRUE,
  warning = TRUE
)

# Core libs
library(rethinking)
library(posterior)
library(cmdstanr)
library(dplyr)
library(tidyr)
library(ggplot2)
library(ggridges)
library(readr)
library(stringr)
library(purrr)
```


```{r setup, include=FALSE}
# If you rely on helper functions (e.g., make_year_dist_mat), source them here:
source("./stir-bayes-backend.R")
model_version <- if (exists("model_version")) model_version else "v1p8"
```

```{r helpers-nonneg-email, include=FALSE}
# ------------------------------------------------------------------
# Helper utilities for physically admissible predictions and optional
# sampling-complete email notification.
#
# Rationale for non-negativity helper:
# v1p8 is intentionally left on the original Gaussian modeling scale to
# preserve the better sampler behavior we observed relative to v1p9.
# To avoid impossible reported predictions, we project concentration and
# volume predictions onto the feasible domain only at prediction time.
# ------------------------------------------------------------------
clamp_nonnegative <- function(x) {
  x <- as.numeric(x)
  x[is.finite(x)] <- pmax(x[is.finite(x)], 0)
  x
}

# Optional email notification after long sampling runs.
# To enable, define these environment variables before knitting/running:
#   STAN_NOTIFY_TO
#   STAN_NOTIFY_FROM
#   STAN_NOTIFY_USER
#   STAN_NOTIFY_PASS
#   STAN_NOTIFY_HOST   (for example: smtp.gmail.com)
#   STAN_NOTIFY_PORT   (for example: 465 or 587)
# Safe to leave in place when SMTP credentials are not set.
notify_sampling_done <- function(subject = NULL, body = NULL) {
  required <- c(
    "STAN_NOTIFY_TO",
    "STAN_NOTIFY_FROM",
    "STAN_NOTIFY_USER",
    "STAN_NOTIFY_PASS",
    "STAN_NOTIFY_HOST",
    "STAN_NOTIFY_PORT"
  )

  vals <- Sys.getenv(required, unset = "")
  names(vals) <- required

  if (any(vals == "")) {
    missing_vars <- names(vals)[vals == ""]
    message(
      "Sampling-finished email skipped: missing environment variable(s): ",
      paste(missing_vars, collapse = ", ")
    )
    return(invisible(FALSE))
  }

  if (!requireNamespace("blastula", quietly = TRUE)) {
    message("Sampling-finished email skipped: install.packages('blastula') is required.")
    return(invisible(FALSE))
  }

  smtp_port <- suppressWarnings(as.integer(Sys.getenv("STAN_NOTIFY_PORT")))
  if (is.na(smtp_port)) {
    message("Sampling-finished email skipped: STAN_NOTIFY_PORT is not a valid integer.")
    return(invisible(FALSE))
  }

  use_ssl <- identical(smtp_port, 465L)

  if (is.null(subject)) {
    subject <- sprintf("Stan sampling finished: %s", model_version)
  }

  if (is.null(body)) {
    body <- paste0(
      "Your CmdStan fit finished sampling.

",
      "Model version: ", model_version, "
",
      "Time finished: ", format(Sys.time(), "%Y-%m-%d %H:%M:%S %Z")
    )
  }

  email_obj <- blastula::compose_email(
    body = blastula::md(body)
  )

  tryCatch({
    blastula::smtp_send(
      email = email_obj,
      from = Sys.getenv("STAN_NOTIFY_FROM"),
      to = Sys.getenv("STAN_NOTIFY_TO"),
      subject = subject,
      credentials = blastula::creds_envvar(
        user = Sys.getenv("STAN_NOTIFY_USER"),
        pass_envvar = "STAN_NOTIFY_PASS",
        host = Sys.getenv("STAN_NOTIFY_HOST"),
        port = smtp_port,
        use_ssl = use_ssl
      )
    )

    message("Sampling-finished email sent.")
    invisible(TRUE)
  }, error = function(e) {
    message("Sampling-finished email failed: ", conditionMessage(e))
    invisible(FALSE)
  })
}
```

# Model 1.8 

* Joint Bayesian model for **OUT concentration and OUT volume** fit simultaneously
* Model uses multivariate normal (MVN) hierarchical structure for analyte-specific parameters
* Model is parametrized in **non-centered** form for better sampling efficiency
* **Analyte-specific** intercepts, STIR slopes, and inflow-concentration slopes
* **Volume** modeled as a function of STIR, with Bayesian imputation for missing VOL
* **Inflow concentration (CIN)** now handled via Bayesian imputation (`CIN ~ dnorm(0,1)`), not zero-substitution
* Includes all DAG-identified adjustment variables affecting concentration
  * { Analyte, Block, Inflow Conc., Irrigation Count, Outflow vol, Sampler Method, duplicate status, flume type }
* Full uncertainty propagation from CIN, VOL, STIR into analyte-specific concentration predictions
* Supports **scenario-based simulation** for STIR → Load via generative functions (conditional or averaged scenarios)
* Incorporates multi-output Gaussian Process (GP) on Year to capture temporal correlation in residuals for each analyte
* v1.5 only has cmdstan model, no rethinking/ulam version
* v1.6 adds crop type effects and residue cover at planting effects
* v1.7 promotes residue, volume, and inflow concentration to latent processes with full Bayesian imputation and joint propagation. Also uses new python wq_cleaned.csv, which has been tested to be identical to previous versions.


# 1. Load and lightly type-enforce the Bayes-clean dataset (from Python pipeline)

```{r load-clean}
# Prefer the Python-produced Bayes-clean dataset
clean_path <- "out/wq_cleaned.csv"
if (!file.exists(clean_path)) {
  alt <- file.path("..", clean_path)
  if (file.exists(alt)) {
    clean_path <- alt
  } else {
    stop(
      "Bayes-clean file not found. Expected 'out/wq_cleaned.csv'.
",
      "Run the Python pipeline first:
",
      "  python code/run_pipeline.py
"
    )
  }
}

wq_clean <- readr::read_csv(clean_path, show_col_types = FALSE)

abbr_levels <- c("NH4","ICP","NO3","NOx","NO2","NPOC","OP","Se","TDS","TKN","TN","TP","TSP","TSS")

# Defensive type enforcement (lightweight, no re-cleaning)
wq_clean <- wq_clean %>%
  dplyr::mutate(
    # Dates
    Date        = if ("Date" %in% names(.)) as.Date(Date) else Date,
    PlantDate   = if ("PlantDate" %in% names(.)) as.Date(PlantDate) else PlantDate,
    HarvestDate = if ("HarvestDate" %in% names(.)) as.Date(HarvestDate) else HarvestDate,

    # Numeric/integer fields commonly used downstream
    Year        = if ("Year" %in% names(.)) as.integer(Year) else Year,
    SeasonYear  = if ("SeasonYear" %in% names(.)) as.integer(SeasonYear) else SeasonYear,

    # Factors/categoricals often used downstream
    Treatment    = if ("Treatment" %in% names(.)) factor(Treatment, levels = c("CT","MT","ST")) else Treatment,
    Rep          = if ("Rep" %in% names(.)) factor(Rep) else Rep,
    Irrigation   = if ("Irrigation" %in% names(.)) factor(Irrigation) else Irrigation,
    SampleMethod = if ("SampleMethod" %in% names(.)) factor(SampleMethod) else SampleMethod,
    FlumeMethod  = if ("FlumeMethod" %in% names(.)) factor(FlumeMethod) else FlumeMethod,
    Duplicate    = if ("Duplicate" %in% names(.)) as.integer(Duplicate) else Duplicate,
    Analyte      = if ("Analyte" %in% names(.)) factor(Analyte) else Analyte,
    analyte_abbr = if ("analyte_abbr" %in% names(.)) factor(analyte_abbr, levels = abbr_levels) else analyte_abbr,
    Crop         = if ("Crop" %in% names(.)) factor(Crop) else Crop,
    previous_crop = if ("previous_crop" %in% names(.)) factor(previous_crop) else NA,

    # Add original row number for tracking
    orig_row = dplyr::row_number()
  )


# If previous_crop was not provided by the pipeline, derive it as last year's Crop
# within plot (Treatment × Rep), and for the first observed year, set previous_crop = Crop.
if (!("previous_crop" %in% names(wq_clean)) || all(is.na(wq_clean$previous_crop))) {
  wq_clean <- wq_clean %>%
    group_by(Treatment, Rep) %>%
    arrange(Year, .by_group = TRUE) %>%
    mutate(previous_crop = dplyr::lag(Crop)) %>%
    mutate(previous_crop = ifelse(is.na(previous_crop), as.character(Crop), as.character(previous_crop))) %>%
    ungroup() %>%
    mutate(previous_crop = factor(previous_crop, levels = levels(factor(Crop))))
}

# Sanity checks (quick)
req_cols <- c("cout_z","cin_z","stir_season_z","volume_z","Treatment","Rep","SampleMethod","MeasureMethod","FlumeMethod","Irrigation","Duplicate","Crop","previous_crop","analyte_abbr","Result_mg_L","RL_mg_L")
missing_cols <- setdiff(req_cols, names(wq_clean))
if (length(missing_cols) > 0) {
  warning("Missing expected columns in wq_cleaned.csv: ", paste(missing_cols, collapse = ", "))
}

cat("Loaded Bayes-clean dataset:", clean_path, "
")
cat("Rows:", nrow(wq_clean), " | Columns:", ncol(wq_clean), "
")

dplyr::glimpse(wq_clean)
```

---

# 2. Prepare model-ready dataset

Select analytes for modeling:

```{r subset-analytes}
# analytes_keep <- c(
#   "OP", "TP", "ICP", "NO3", "TN", "TSS",
#   "TKN", "TSP", "NH4", "NPOC", "Se", "NO2",
#   "NOx", "TDS"
# )

analytes_keep <- c(
  "OP", "TP", "NO3", "TN", "TSS",
  "TKN", "NH4", "Se", "NO2", "TDS"
)

d_mod <- wq_clean %>%
  dplyr::filter(analyte_abbr %in% analytes_keep) %>%
  dplyr::mutate(mod_row = dplyr::row_number()) %>%
  droplevels()

na_table <- d_mod %>%
  dplyr::group_by(analyte_abbr) %>%
  dplyr::summarize(
    n_total   = dplyr::n(),
    n_value   = sum(!is.na(Result_mg_L)),
    n_na      = sum(is.na(Result_mg_L)),
    prop_na   = n_na / n_total, #proportion missing
    .groups   = "drop"
  )

na_table

```

Create analyte index and build data list for ulam:

```{r make-stan-dat}
# 2. Build indices and standardized predictors -----------------------------
# DROP-IN replacement (v1p8, corrected):
# - Ensures censored rows are NOT treated as missing in C by constructing C_vec (dummy-filled on censored rows)
# - Removes PrevCr_n from stan_dat (Stan uses Cr_n for both Cr and PrevCr indexing)
# - Adds explicit guardrails that catch the main failure modes before you run Stan

residue_metric <- "cover"  # "cover" or "drymass"

# ---- Required pipeline columns / fallbacks ----
vin_candidates <- c("vin_z","inflow_volume_z","inflow_vol_z","vin_std")
vin_col <- vin_candidates[vin_candidates %in% names(d_mod)][1]
if (is.na(vin_col)) {
  stop("No inflow-volume z-score column found. Expected one of: ", paste(vin_candidates, collapse = ", "))
}

irr_z_candidates <- c("irr_z","IRR_z","irr_num_z")
irr_z_col <- irr_z_candidates[irr_z_candidates %in% names(d_mod)][1]
if (is.na(irr_z_col)) {
  stop("No irrigation z-score column found. Expected one of: ", paste(irr_z_candidates, collapse = ", "),
       ". (Pipeline should provide IRR_z in v1p8 workflow.)")
}

# Censoring columns (pipeline-provided)
if (!("Result_is_nd" %in% names(d_mod))) {
  stop("Missing 'Result_is_nd' column (0/1). v1p8 expects pipeline to provide this.")
}
lod_candidates <- c("Result_lod_mg_L","RL_mg_L")
lod_col <- lod_candidates[lod_candidates %in% names(d_mod)][1]
if (is.na(lod_col)) {
  stop("Missing censoring limit column. Expected one of: ", paste(lod_candidates, collapse = ", "))
}

# ---- Mutate indices + covariates ----
d_mod <- d_mod %>%
  mutate(
    # analyte index
    A = as.integer(analyte_abbr),

    # Block (Rep) index
    B = as.integer(Rep),

    # Sampler method index
    S = as.integer(SampleMethod),

    # Flume type index
    Fu = as.integer(FlumeMethod),

    # Irrigation count as numeric (still used for filtering out storms; model uses IRR_z)
    irr_num = dplyr::case_when(
      as.character(Irrigation) %in% c("S1", "S2") ~ NA_integer_,
      TRUE ~ as.integer(readr::parse_number(as.character(Irrigation)))
    ),

    # Inflow concentration (standardized), keep NAs for Bayesian imputation
    cin_std = cin_z,

    # Inflow volume (standardized), keep NAs for Bayesian imputation
    vin_std = .data[[vin_col]],

    # Irrigation z-score (standardized), from pipeline
    irr_z = .data[[irr_z_col]],

    # Duplicate status as 0/1
    dup = as.integer(Duplicate),

    # Year index
    Y = as.integer(as.factor(Year)),

    # Crop type index (categorical)
    Cr = as.integer(Crop),

    # Previous crop type index (categorical, must align to same levels as Crop)
    PrevCr = as.integer(previous_crop),

    # -----------------------------------------------------------------------
    # Residue percent cover -> proportion in (0,1) with epsilon adjustment
    # -----------------------------------------------------------------------
    residue_prop = {
      x <- as.numeric(.data$Residue_PercentCover)
      x <- ifelse(is.na(x), NA_real_, ifelse(x > 1, x / 100, x))
      eps <- 1e-4
      ifelse(is.na(x), NA_real_, pmin(pmax(x, eps), 1 - eps))
    },

    # -----------------------------------------------------------------------
    # Residue dry mass (kg/m^2) -> log transform (<=0 becomes NA)
    # -----------------------------------------------------------------------
    drymass_kg_m2 = as.numeric(.data$Residue_DryMass_kg_m2),
    drymass_log = dplyr::if_else(
      is.na(drymass_kg_m2) | drymass_kg_m2 <= 0,
      NA_real_,
      log(drymass_kg_m2)
    )
  )

# Drop storm rows (no ordinal irrigation meaning)
d_mod <- d_mod %>%
  dplyr::filter(!is.na(irr_num))

# Z-score drymass_log using observed values only (keeps NA as NA)
drymass_mu <- mean(d_mod$drymass_log, na.rm = TRUE)
drymass_sd <- sd(d_mod$drymass_log, na.rm = TRUE)

d_mod <- d_mod %>%
  mutate(
    drymass_log_z = (drymass_log - drymass_mu) / drymass_sd,

    # Single residue input used by Stan
    RES_input = dplyr::case_when(
      residue_metric == "cover"   ~ residue_prop,
      residue_metric == "drymass" ~ drymass_log_z,
      TRUE ~ residue_prop
    )
  )

# ---- Censoring vectors for Stan (C on cout_z scale) ----
C_cens <- as.integer(d_mod$Result_is_nd == 1)
lod_mg_L <- as.numeric(d_mod[[lod_col]])

# Convert mg/L LOD -> cout_z scale by analyte-specific linear mapping:
#   cout_z ≈ alpha_a + beta_a * log(Result_mg_L)  (fit using non-censored observations)
C_cens_limit <- rep(NA_real_, nrow(d_mod))

for (a in sort(unique(d_mod$A))) {
  idx_a <- which(d_mod$A == a)

  fit_idx <- idx_a[
    !is.na(d_mod$cout_z[idx_a]) &
      !is.na(d_mod$Result_mg_L[idx_a]) &
      (d_mod$Result_is_nd[idx_a] == 0) &
      (d_mod$Result_mg_L[idx_a] > 0) &
      !is.na(lod_mg_L[idx_a]) & (lod_mg_L[idx_a] > 0)
  ]

  if (length(fit_idx) >= 20) {
    x <- log(d_mod$Result_mg_L[fit_idx])
    y <- d_mod$cout_z[fit_idx]
    co <- coef(lm(y ~ x))
    alpha <- unname(co[1]); beta <- unname(co[2])
    C_cens_limit[idx_a] <- alpha + beta * log(lod_mg_L[idx_a])
  }
}

# Fallback for analytes with too few points for a stable mapping
if (anyNA(C_cens_limit)) {
  fit_all <- which(
    !is.na(d_mod$cout_z) &
      !is.na(d_mod$Result_mg_L) &
      (d_mod$Result_is_nd == 0) &
      (d_mod$Result_mg_L > 0) &
      !is.na(lod_mg_L) & (lod_mg_L > 0)
  )
  if (length(fit_all) < 50) {
    stop("Not enough non-censored observations to infer cout_z <-> log(mg/L) mapping for censoring limits.")
  }
  co_all <- coef(lm(d_mod$cout_z[fit_all] ~ log(d_mod$Result_mg_L[fit_all])))
  miss <- which(is.na(C_cens_limit))
  C_cens_limit[miss] <- unname(co_all[1]) + unname(co_all[2]) * log(lod_mg_L[miss])
}

# Safety: Stan only uses limit when C_cens==1, but avoid NA/Inf anyway
C_cens_limit[!is.finite(C_cens_limit)] <- 0

# ---- Critical fix: construct C_vec so censored rows are NOT treated as missing ----
# If cout_z is NA on ND rows, we provide a dummy finite value (ignored by Stan when C_cens==1).
C_vec <- d_mod$cout_z
C_vec[C_cens == 1 & is.na(C_vec)] <- 0

# instantiate distance matrix for gaussian process on year
years_used <- sort(unique(d_mod$Year))
mat <- make_year_dist_mat(years_used)

# lookup code for later
analyte_lookup <- d_mod %>%
  distinct(A, analyte_abbr) %>%
  arrange(A)

block_lookup <- d_mod %>%
  distinct(B, Rep) %>%
  arrange(B)

sampler_lookup <- d_mod %>%
  distinct(S, SampleMethod) %>%
  arrange(S)

flume_lookup <- d_mod %>%
  distinct(Fu, FlumeMethod) %>%
  arrange(Fu)

irrigation_lookup <- d_mod %>%
  distinct(irr_num, Irrigation) %>%
  arrange(irr_num)

duplicate_lookup <- d_mod %>%
  distinct(dup, Duplicate) %>%
  arrange(dup)

crop_lookup <- d_mod %>%
  distinct(Cr, Crop) %>%
  arrange(Cr)

prevcrop_lookup <- d_mod %>%
  distinct(PrevCr, previous_crop) %>%
  arrange(PrevCr)

# prepare data list for stan
stan_dat <- list(
  N    = nrow(d_mod),

  # group counts
  A_n  = length(unique(d_mod$A)),
  B_n  = length(unique(d_mod$B)),
  S_n  = length(unique(d_mod$S)),
  F_n  = length(unique(d_mod$Fu)),
  D_n  = length(unique(d_mod$dup)),
  Y_n  = length(unique(d_mod$Y)),
  Cr_n = length(unique(d_mod$Cr)),

  # indices
  A    = d_mod$A,
  B    = d_mod$B,
  S    = d_mod$S,
  Fu   = d_mod$Fu,
  Y    = d_mod$Y,
  Cr   = d_mod$Cr,
  PrevCr = d_mod$PrevCr,

  # distance matrix for year effects
  D    = mat,

  # outcome + censoring (use C_vec, not raw cout_z)
  C    = C_vec,
  C_cens = C_cens,
  C_cens_limit = C_cens_limit,

  # main exposure
  STIR = d_mod$stir_season_z,

  # other covariates from DAG
  CIN   = d_mod$cin_std,
  VOL   = d_mod$volume_z,
  IRR_z = d_mod$irr_z,
  VIN   = d_mod$vin_std,
  DUP   = d_mod$dup,

  # residue covariate (cover or drymass, depending on residue_metric)
  RES  = d_mod$RES_input
)

cat("\nCensoring summary:\n")
cat("  Censored (ND) rows:", sum(stan_dat$C_cens == 1), "\n")
cat("  Censored rows with finite limit:", sum(stan_dat$C_cens == 1 & is.finite(stan_dat$C_cens_limit)), "\n")
cat("  Censored rows with finite C:", sum(stan_dat$C_cens == 1 & is.finite(stan_dat$C)), "\n")

# Optional: quick guardrails
if (identical(residue_metric, "cover")) {
  stopifnot(all(is.na(stan_dat$RES) | (stan_dat$RES > 0 & stan_dat$RES < 1)))
}

# Additional quick checks for censoring inputs
stopifnot(length(stan_dat$C_cens) == stan_dat$N)
stopifnot(length(stan_dat$C_cens_limit) == stan_dat$N)
stopifnot(all(stan_dat$C_cens %in% c(0L, 1L)))

# Extra guardrails that specifically protect censoring correctness
if (any(C_cens == 1 & is.na(d_mod$cout_z))) {
  message("Note: cout_z is NA for some censored rows; using dummy C=0 for those rows so censoring likelihood is applied.")
}
stopifnot(all(is.finite(stan_dat$C[C_cens == 1])))
stopifnot(all(is.finite(stan_dat$C_cens_limit[C_cens == 1])))
#stopifnot(all(stan_dat$C_cens_limit[stan_dat$C_cens == 1] != 0))

```

```{r check-na-vars}
# Identify the stan_dat entries that are vectors of length N (i.e., row-level variables)
vec_names <- names(stan_dat)[sapply(stan_dat, function(x) length(x) == stan_dat$N)]

# Loop through these and report NAs
for (nm in vec_names) {
  vals <- stan_dat[[nm]]
  na_idx <- which(is.na(vals))

  cat("\n==============================\n")
  cat("Variable:", nm, "\n")

  if (length(na_idx) == 0) {
    cat("No NA values detected.\n")
  } else {
    cat("NA count:", length(na_idx), "\n")
    cat("Rows with NA:", paste(head(na_idx, 25), collapse = ", "),
        if (length(na_idx) > 25) "..." else "", "\n\n")

    # Show the corresponding d_mod rows
    cat("Example offending rows from d_mod:\n")
    print(d_mod[na_idx[1:min(10, length(na_idx))], ])
  }
}

```


# 3. Hierarchical Bayesian model

```{r cmdstan model fit}
# ======================================================================
# Fit cmdstan model (v1p8) with censoring + missingness (CIN/VIN/VOL/RES)
# CLEAN drop-in replacement for your "Fit cmdstan model" block
#
# Key points:
#   1) ND rows are handled as LEFT-CENSORED (do NOT mark them missing).
#   2) Truly missing C rows (and not censored) are handled via missidx.
#   3) Stan receives NO NA values in any vector.
#   4) Censoring limits are transformed onto the same scale as cout_z
#      using an analyte-specific mapping: cout_z ≈ alpha_a + beta_a*log(mg/L).
# ======================================================================

# Force recompile to ensure v1p8 is what runs
mod <- cmdstan_model("m_stir_mogp_v1p8.stan", force_recompile = TRUE)

# Start from base stan_dat you built earlier
stan_dat_cmd <- stan_dat

# -------------------------------------------------------------------
# 1) Pull row-level vectors (length N)
# -------------------------------------------------------------------
N <- stan_dat_cmd$N

C_raw   <- d_mod$cout_z
CIN_vec <- d_mod$cin_std
VIN_vec <- d_mod$vin_std
VOL_vec <- d_mod$volume_z

Cr_vec    <- d_mod$Cr
PrevCr_vec <- d_mod$PrevCr

# Residue input, use the same one you passed in stan_dat
# (RES_input is preferred because it already reflects residue_metric choice)
if ("RES_input" %in% names(d_mod)) {
  RES_vec_raw <- d_mod$RES_input
} else if ("residue_prop" %in% names(d_mod)) {
  RES_vec_raw <- d_mod$residue_prop
} else if ("Residue_PercentCover" %in% names(d_mod)) {
  RES_vec_raw <- as.numeric(d_mod$Residue_PercentCover)
  RES_vec_raw <- ifelse(is.na(RES_vec_raw), NA_real_,
                        ifelse(RES_vec_raw > 1, RES_vec_raw / 100, RES_vec_raw))
} else {
  stop("No residue column found. Expected one of: RES_input, residue_prop, Residue_PercentCover.")
}

stopifnot(
  length(C_raw)       == N,
  length(CIN_vec)     == N,
  length(VIN_vec)     == N,
  length(VOL_vec)     == N,
  length(Cr_vec)      == N,
  length(PrevCr_vec)  == N,
  length(RES_vec_raw) == N
)

# -------------------------------------------------------------------
# 2) Censoring: ND flag + RL/LOD, convert RL to cout_z scale
# -------------------------------------------------------------------
if (!("Result_is_nd" %in% names(d_mod))) {
  stop("Expected pipeline column 'Result_is_nd' (0/1) for v1p8 censoring.")
}
C_cens <- as.integer(d_mod$Result_is_nd == 1)

lod_candidates <- c("Result_lod_mg_L", "RL_mg_L")
lod_col <- lod_candidates[lod_candidates %in% names(d_mod)][1]
if (is.na(lod_col)) stop("Need 'Result_lod_mg_L' or 'RL_mg_L' to build censor limits.")
lod_mg_L <- as.numeric(d_mod[[lod_col]])

# Convert mg/L RL -> cout_z using analyte-specific mapping:
#   cout_z ≈ alpha_a + beta_a * log(Result_mg_L)
C_cens_limit <- rep(NA_real_, N)

for (a in sort(unique(d_mod$A))) {
  idx_a <- which(d_mod$A == a)

  fit_idx <- idx_a[
    (C_cens[idx_a] == 0L) &
      !is.na(C_raw[idx_a]) &
      !is.na(d_mod$Result_mg_L[idx_a]) & (d_mod$Result_mg_L[idx_a] > 0) &
      !is.na(lod_mg_L[idx_a]) & (lod_mg_L[idx_a] > 0)
  ]

  if (length(fit_idx) >= 20) {
    x <- log(d_mod$Result_mg_L[fit_idx])
    y <- C_raw[fit_idx]
    co <- coef(lm(y ~ x))
    alpha <- unname(co[1]); beta <- unname(co[2])
    C_cens_limit[idx_a] <- alpha + beta * log(lod_mg_L[idx_a])
  }
}

# Global fallback for any remaining limits
if (anyNA(C_cens_limit)) {
  fit_all <- which(
    (C_cens == 0L) &
      !is.na(C_raw) &
      !is.na(d_mod$Result_mg_L) & (d_mod$Result_mg_L > 0) &
      !is.na(lod_mg_L) & (lod_mg_L > 0)
  )
  if (length(fit_all) < 50) stop("Too few detected observations to infer censoring limit transform.")
  co_all <- coef(lm(C_raw[fit_all] ~ log(d_mod$Result_mg_L[fit_all])))

  miss <- which(is.na(C_cens_limit))
  C_cens_limit[miss] <- unname(co_all[1]) + unname(co_all[2]) * log(lod_mg_L[miss])
}

# Safety: censor limits must be finite for censored rows; non-censored rows can be anything finite
C_cens_limit[!is.finite(C_cens_limit)] <- 0

# -------------------------------------------------------------------
# 3) Missingness for C: DO NOT drop censored rows
# -------------------------------------------------------------------
# Missing if C is NA AND NOT censored
is_C_miss <- as.integer(is.na(C_raw) & (C_cens == 0L))
C_missidx <- which(is_C_miss == 1L)

# Stan-safe C:
# - observed values stay
# - censored rows with NA get dummy finite value (ignored by Stan because C_cens==1)
# - true-missing rows filled with placeholder
C_clean <- as.numeric(C_raw)
C_clean[is.na(C_clean) & (C_cens == 1L)] <- 0
C_clean[is.na(C_clean) & (C_cens == 0L)] <- 0

# Guardrail: no censored row should be marked missing
stopifnot(all(is_C_miss[C_cens == 1L] == 0L))
stopifnot(all(is.finite(C_clean[C_cens == 1L])))
stopifnot(all(is.finite(C_cens_limit[C_cens == 1L])))

# -------------------------------------------------------------------
# 4) Helpers for other missings (CIN/VIN/VOL/RES)
# -------------------------------------------------------------------
prep_miss_gaussian <- function(x, fill = 0) {
  missidx <- which(is.na(x))
  x_clean <- as.numeric(x)
  if (length(missidx) > 0) x_clean[missidx] <- fill
  list(missidx = missidx, N_miss = length(missidx), clean = x_clean)
}

prep_miss_residue <- function(x, eps = 1e-4, fill = 0.5) {
  x_num <- as.numeric(x)
  missidx <- which(is.na(x_num))
  obs_idx <- setdiff(seq_along(x_num), missidx)

  if (length(obs_idx) > 0) {
    bad_nonfinite <- obs_idx[!is.finite(x_num[obs_idx])]
    if (length(bad_nonfinite) > 0) {
      stop("RES contains non-finite values at rows: ",
           paste(head(bad_nonfinite, 20), collapse = ", "),
           if (length(bad_nonfinite) > 20) " ..." else "")
    }
  }

  # If user accidentally passed 0–100 here, convert to 0–1
  if (length(obs_idx) > 0 && any(x_num[obs_idx] > 1, na.rm = TRUE)) {
    x_num[obs_idx] <- x_num[obs_idx] / 100
  }

  if (length(obs_idx) > 0) {
    bad_range <- obs_idx[x_num[obs_idx] < 0 | x_num[obs_idx] > 1]
    if (length(bad_range) > 0) {
      stop("RES has values outside [0,1] at rows: ",
           paste(head(bad_range, 20), collapse = ", "),
           if (length(bad_range) > 20) " ..." else "",
           "\nCheck residue scaling upstream (should be a 0–1 proportion).")
    }
  }

  if (length(obs_idx) > 0) {
    x_num[obs_idx] <- pmin(pmax(x_num[obs_idx], eps), 1 - eps)
  }

  if (length(missidx) > 0) x_num[missidx] <- fill

  list(missidx = missidx, N_miss = length(missidx), clean = x_num)
}

imp_CIN <- prep_miss_gaussian(CIN_vec, fill = 0)
imp_VIN <- prep_miss_gaussian(VIN_vec, fill = 0)
imp_VOL <- prep_miss_gaussian(VOL_vec, fill = 0)
imp_RES <- prep_miss_residue(RES_vec_raw, eps = 1e-4, fill = 0.5)

# -------------------------------------------------------------------
# 5) Populate cmdstan data list (NO NA allowed)
# -------------------------------------------------------------------
# Concentration
stan_dat_cmd$C            <- C_clean
stan_dat_cmd$C_cens       <- as.array(as.integer(C_cens))
stan_dat_cmd$C_cens_limit <- as.numeric(C_cens_limit)

stan_dat_cmd$N_C_miss     <- length(C_missidx)
stan_dat_cmd$C_missidx    <- if (length(C_missidx) == 0) integer(0) else as.array(as.integer(C_missidx))
stan_dat_cmd$is_C_miss    <- as.array(as.integer(is_C_miss))

# CIN
stan_dat_cmd$CIN         <- imp_CIN$clean
stan_dat_cmd$N_CIN_miss  <- imp_CIN$N_miss
stan_dat_cmd$CIN_missidx <- if (imp_CIN$N_miss == 0) integer(0) else as.array(as.integer(imp_CIN$missidx))

# VIN
stan_dat_cmd$VIN         <- imp_VIN$clean
stan_dat_cmd$N_VIN_miss  <- imp_VIN$N_miss
stan_dat_cmd$VIN_missidx <- if (imp_VIN$N_miss == 0) integer(0) else as.array(as.integer(imp_VIN$missidx))

# VOL
stan_dat_cmd$VOL         <- imp_VOL$clean
stan_dat_cmd$N_VOL_miss  <- imp_VOL$N_miss
stan_dat_cmd$VOL_missidx <- if (imp_VOL$N_miss == 0) integer(0) else as.array(as.integer(imp_VOL$missidx))
stan_dat_cmd$is_VOL_miss <- as.array(as.integer(is.na(VOL_vec)))

# RES
stan_dat_cmd$RES         <- imp_RES$clean
stan_dat_cmd$N_RES_miss  <- imp_RES$N_miss
stan_dat_cmd$RES_missidx <- if (imp_RES$N_miss == 0) integer(0) else as.array(as.integer(imp_RES$missidx))
stan_dat_cmd$is_RES_miss <- as.array(as.integer(is.na(RES_vec_raw)))

# Crop indices (already integer-coded upstream)
Cr_n_now <- length(unique(Cr_vec))
stan_dat_cmd$Cr <- as.array(as.integer(Cr_vec))
if (is.null(stan_dat_cmd$Cr_n) || stan_dat_cmd$Cr_n != Cr_n_now) {
  stan_dat_cmd$Cr_n <- Cr_n_now
}

# Previous crop index (must be within 1..Cr_n as declared in Stan)
stan_dat_cmd$PrevCr <- as.array(as.integer(PrevCr_vec))

# -------------------------------------------------------------------
# 6) Final consistency checks + helpful diagnostics
# -------------------------------------------------------------------
stopifnot(
  all(is.finite(stan_dat_cmd$C)),
  all(is.finite(stan_dat_cmd$CIN)),
  all(is.finite(stan_dat_cmd$VIN)),
  all(is.finite(stan_dat_cmd$VOL)),
  all(is.finite(stan_dat_cmd$RES)),
  all(stan_dat_cmd$RES > 0),
  all(stan_dat_cmd$RES < 1),
  all(stan_dat_cmd$C_cens %in% c(0L, 1L)),
  length(stan_dat_cmd$C_cens) == N,
  length(stan_dat_cmd$C_cens_limit) == N,
  isTRUE(all(stan_dat_cmd$Cr >= 1L)),
  isTRUE(all(stan_dat_cmd$Cr <= stan_dat_cmd$Cr_n)),
  isTRUE(all(stan_dat_cmd$PrevCr >= 1L)),
  isTRUE(all(stan_dat_cmd$PrevCr <= stan_dat_cmd$Cr_n))
)

cat("\n[INFO] Missing counts:\n")
cat("  N_C_miss    =", stan_dat_cmd$N_C_miss, "\n")
cat("  N_CIN_miss  =", stan_dat_cmd$N_CIN_miss, "\n")
cat("  N_VIN_miss  =", stan_dat_cmd$N_VIN_miss, "\n")
cat("  N_VOL_miss  =", stan_dat_cmd$N_VOL_miss, "\n")
cat("  N_RES_miss  =", stan_dat_cmd$N_RES_miss, "\n")
cat("[INFO] Crop levels (Cr_n) =", stan_dat_cmd$Cr_n, "\n")

cat("\n[INFO] Censoring summary:\n")
cat("  Censored (ND) rows:", sum(stan_dat_cmd$C_cens == 1L), "\n")
cat("  Censored rows with finite limit:",
    sum(stan_dat_cmd$C_cens == 1L & is.finite(stan_dat_cmd$C_cens_limit)), "\n")
cat("  Censored rows marked missing (should be 0):",
    sum(stan_dat_cmd$C_cens == 1L & stan_dat_cmd$is_C_miss == 1L), "\n")

cat("\n[INFO] Residue summary (raw, before clamp/fill):\n")
print(summary(RES_vec_raw))
cat("[INFO] Residue summary (Stan-ready, after clamp/fill):\n")
print(summary(stan_dat_cmd$RES))

# -------------------------------------------------------------------
# 7) Sample with cmdstanr
# -------------------------------------------------------------------
dir.create("out_cmdstanr", showWarnings = FALSE)

fit <- mod$sample(
  data            = stan_dat_cmd,
  chains          = 4,
  parallel_chains = 4,
  iter_warmup     = 2000,
  iter_sampling   = 1000,
  refresh         = 100,
  show_messages   = TRUE,
  output_dir      = "out_cmdstanr",
  # save_warmup     = TRUE,
  adapt_delta     = 0.95,
  init            = 0.1,
  max_treedepth   = 15
)

```


```{r cmdstan model fit}
fit_rds_path <- "./out_cmdstanr/fit_mogp_v1p8.rds"
saveRDS(fit, file = fit_rds_path)
# fit <- readRDS(fit_rds_path)

# Optional email notification after fit serialization.
# This is intentionally non-fatal so overnight runs are not interrupted if
# SMTP credentials are missing or Gmail rejects the message.
notify_sampling_done(
  subject = sprintf("Stan sampling finished: %s", model_version),
  body = paste0(
    "Your CmdStan fit finished sampling.

",
    "Model version: ", model_version, "
",
    "Saved fit: ", normalizePath(fit_rds_path, winslash = "/", mustWork = FALSE), "
",
    "Time finished: ", format(Sys.time(), "%Y-%m-%d %H:%M:%S %Z")
  )
)

```


---

# 4. Posterior inspection

```{r}
# ============================================================
# v1p8 DIAGNOSTICS: key (non-latent) research parameters
# - Produces a clean table of R-hat + ESS (bulk/tail)
# - Uses robust regex-based inclusion/exclusion so it won't
#   break when parameter names change (e.g., gamma_CR removed)
#
# Assumes you have a cmdstanr fit object named: fit
# ============================================================

# --- Helper: get all variable names present in the fit ---
.all_vars <- function(fit) {
  # posterior::variables() works on draws objects
  draws <- fit$draws()
  posterior::variables(draws)
}

# --- Helper: filter to "important" parameters ---
pick_key_vars_v1p8 <- function(fit) {
  vars <- .all_vars(fit)

  # 1) Explicit INCLUDE patterns (edit freely)
  #    These are meant to capture your research-facing parameters:
  include_patterns <- c(
    # Global / analyte-level regression coefficients (common naming styles)
    "^alpha", "^beta", "^b_", "^a_",

    # GP hyperparameters (often key to year structure and mixing)
    "etasq", "rhosq", "rho", "eta", "sigma_year", "l_year", "ls_year",

    # Correlation / covariance structures (if you have multi-output components)
    "Omega", "L_Omega", "Sigma", "L_Sigma", "cholesky", "corr",

    # Observation / process SDs that often drive funnels & divergences
    "^sigma", "tau"
  )

  # 2) Explicit EXCLUDE patterns (these remove latent / per-row stuff)
  #    This is where you keep the list aggressive so you only see “research” params.
  exclude_patterns <- c(
    # latent truths & per-observation latent states
    "C_true", "V_true", "VOL_rep", "CIN_merge", "VIN_merge",
    "mu_C", "mu_V", "mu_res", "RES_star",

    # raw latent z vectors / imputation internals
    "^z_", "zC", "zV", "z_C", "z_V", "C_miss", "CIN_miss", "VIN_miss", "VOL_miss", "RES_miss",

    # posterior predictive per-row
    "_rep$", "rep\\[", "y_rep", "C_rep", "V_rep", "VOL_rep",

    # per-row log-lik arrays (if present)
    "log_lik", "lp__"
  )

  inc_re <- paste(include_patterns, collapse = "|")
  exc_re <- paste(exclude_patterns, collapse = "|")

  key_vars <- vars %>%
    # include anything that matches include patterns
    .[str_detect(., inc_re)] %>%
    # exclude latent / row-level / loglik etc
    .[!str_detect(., exc_re)] %>%
    # drop transformed parameters that are huge arrays unless you want them
    unique()

  # If you want to *force include* some known v1p8 names, add them here:
  # (These are optional; only kept if they actually exist in the fit.)
  force_include <- c(
    # Typical names in your lineage; safe because we intersect with vars
    "beta_stir", "beta_STIR", "b_stir",
    "beta_cin",  "beta_CIN",  "b_cin",
    "beta_vin",  "beta_VIN",  "b_vin",
    "beta_irr",  "beta_IRR",  "b_irr",
    "beta_dup",  "beta_DUP",  "b_dup",
    "beta_res",  "beta_RES",  "b_res",
    "beta_prevcr", "beta_PrevCr", "b_prevcr",
    "etasq_year", "rhosq_year"
  )
  key_vars <- unique(c(key_vars, intersect(force_include, vars)))

  key_vars
}

# --- Main: summarise key vars (Rhat + ESS) ---
summarise_key_vars <- function(fit, max_print = 250) {
  key_vars <- pick_key_vars_v1p8(fit)

  if (length(key_vars) == 0) {
    stop("No key variables matched. Consider relaxing include/exclude patterns in pick_key_vars_v1p8().")
  }

  summ <- fit$summary(variables = key_vars) %>%
    as_tibble() %>%
    # cmdstanr returns columns: variable, mean, median, sd, mad, q5, q95, rhat, ess_bulk, ess_tail
    arrange(desc(rhat), ess_bulk) %>%
    mutate(
      flag_rhat = if_else(is.finite(rhat) & rhat > 1.01, "CHECK", ""),
      flag_ess  = if_else(is.finite(ess_bulk) & ess_bulk < 200, "LOW_ESS", "")
    )

  cat("\n============================================================\n")
  cat("[v1p8] Key parameter diagnostics (sorted by worst R-hat)\n")
  cat("  Total key vars:", nrow(summ), "\n")
  cat("  R-hat > 1.01:", sum(summ$rhat > 1.01, na.rm = TRUE), "\n")
  cat("  ESS bulk < 200:", sum(summ$ess_bulk < 200, na.rm = TRUE), "\n")
  cat("============================================================\n\n")

  print(
    summ %>%
      select(variable, mean, sd, q5, q95, rhat, ess_bulk, ess_tail, flag_rhat, flag_ess) %>%
      head(max_print),
    n = min(max_print, nrow(summ))
  )

  invisible(summ)
}

# --- Optional: per-chain quick check for the worst offenders ---
trace_worst <- function(fit, summ_tbl, top_n = 12) {
  library(bayesplot)
  worst <- summ_tbl %>%
    filter(is.finite(rhat)) %>%
    arrange(desc(rhat), ess_bulk) %>%
    slice_head(n = top_n) %>%
    pull(variable)

  draws_array <- as_draws_array(fit$draws(variables = worst))

  # Trace + rank plots are the fastest way to see “one chain stuck”
  bayesplot::mcmc_trace(draws_array, pars = worst)
  bayesplot::mcmc_rank_overlay(draws_array, pars = worst)
}

# ============================================================
# RUN
# ============================================================
key_summary <- summarise_key_vars(fit, max_print = 400)

# If you want the quick plots for the worst parameters:
trace_worst(fit, key_summary, top_n = 12)
```

```{r}
# If you want to dig into specific parameters that are showing issues, you can pull their draws and summarise by chain:
bad_pars <- c("sigma_C_obs[2]","sigma_C_obs[6]","sigma_C_obs[7]","sigma_analyte","tau_log_sigma_C_obs")

draws <- as_draws_df(fit$draws(variables = bad_pars))

draws %>%
  tidyr::pivot_longer(cols = all_of(bad_pars), names_to = "param", values_to = "value") %>%
  group_by(param, .chain) %>%
  summarise(
    median = median(value),
    q25 = quantile(value, 0.25),
    q75 = quantile(value, 0.75),
    .groups = "drop"
  ) %>%
  arrange(param, .chain) %>%
  print(n = Inf)
```


```{r, fig.width=10, fig.height=7}
# Takes a long time to run (2-3+ hours)
#cmdstan_dashboard(fit, vars = vars_important, trank = TRUE)
cmdstan_dashboard(fit)

# Save a static dashboard image for later review outside the knitted HTML.
# This is non-fatal because some graphics devices or cmdstanr versions may
# behave differently across machines.
dashboard_png <- file.path("out_cmdstanr", paste0("cmdstan_dashboard_", model_version, ".png"))
tryCatch({
  grDevices::png(filename = dashboard_png, width = 2400, height = 1800, res = 220)
  cmdstan_dashboard(fit)
  grDevices::dev.off()
  message("Saved cmdstan dashboard PNG to: ", dashboard_png)
}, error = function(e) {
  if (grDevices::dev.cur() > 1) {
    try(grDevices::dev.off(), silent = TRUE)
  }
  message("Could not save cmdstan dashboard PNG: ", conditionMessage(e))
})

```



```{r stan analyte effects}
## =====================================================================
## Extract posterior samples from cmdstanr into a "rethinking-style" list
## DROP-IN replacement for v1p8 (NO fit$summary; uses fit$draws only)
##
## Core fix vs your failing attempt:
##   cmdstanr/posterior store indexed names like alpha[1], not base "alpha",
##   so we must request indexed variable names explicitly.
##
## v1p8 specifics:
##   - Residue crop effect is by *previous crop*: gamma_PCr_res, sigma_PCr_res
##     (we alias to gamma_Cr_res / sigma_Cr_res for downstream compatibility)
## =====================================================================

dat_used <- if (exists("stan_dat_cmd")) stan_dat_cmd else stan_dat

# ---- Dimensions ----
A_n   <- dat_used$A_n
B_n   <- dat_used$B_n
S_n   <- dat_used$S_n
F_n   <- dat_used$F_n
Y_n   <- dat_used$Y_n
Cr_n  <- dat_used$Cr_n
N     <- dat_used$N
PCr_n <- if (!is.null(dat_used$PrevCr_n)) dat_used$PrevCr_n else Cr_n  # v1p8 residue groups

# Residue placeholder check (all-constant means no supported effect yet)
res_sd <- suppressWarnings(sd(dat_used$RES, na.rm = TRUE))
res_is_placeholder <- is.na(res_sd) || res_sd == 0

# ---- Helpers to build CmdStan variable names (indexed) ----
vnames_vec <- function(name, n) {
  paste0(name, "[", seq_len(n), "]")
}
vnames_mat <- function(name, n1, n2) {
  # CmdStan column order is typically: [1,1],[2,1],...,[n1,1],[1,2],...,[n1,n2]
  out <- character(n1 * n2)
  k <- 1L
  for (j in seq_len(n2)) {
    for (i in seq_len(n1)) {
      out[k] <- paste0(name, "[", i, ",", j, "]")
      k <- k + 1L
    }
  }
  out
}
vnames_3d <- function(name, n1, n2, n3) {
  # Typical order: i fastest, then j, then k
  out <- character(n1 * n2 * n3)
  k <- 1L
  for (kk in seq_len(n3)) {
    for (jj in seq_len(n2)) {
      for (ii in seq_len(n1)) {
        out[k] <- paste0(name, "[", ii, ",", jj, ",", kk, "]")
        k <- k + 1L
      }
    }
  }
  out
}
vnames_2d_y_a <- function(name, Y_n, A_n) {
  # F_year is declared as matrix[Y_n, A_n]
  vnames_mat(name, Y_n, A_n)
}

# ---- Build variable list to pull (indexed + scalars) ----
vars_indexed <- c(
  # analyte-level effects (vectors length A_n)
  vnames_vec("alpha",      A_n),
  vnames_vec("beta_stir",  A_n),
  vnames_vec("beta_cin",   A_n),
  vnames_vec("beta_dup",   A_n),
  vnames_vec("beta_res_C", A_n),

  # random effects (matrices)
  vnames_mat("gamma_B", A_n, B_n),
  vnames_mat("gamma_S", A_n, S_n),
  vnames_mat("gamma_F", A_n, F_n),

  # crop effects
  vnames_mat("gamma_Cr", A_n, Cr_n),
  vnames_vec("gamma_Cr_V", Cr_n),

  # irrigation effect by analyte (vector)
  vnames_vec("beta_irr", A_n),

  # GP year effect: matrix[Y_n, A_n]
  vnames_2d_y_a("F_year", Y_n, A_n),

  # multi-output GP Cholesky correlation matrix (A_n x A_n)
  vnames_mat("L_corr_Agp", A_n, A_n),

  # multi-output GP SDs (vector)
  vnames_vec("sigma_Agp", A_n),

  # concentration SD structure
  vnames_vec("sigma_C_obs", A_n),
  vnames_vec("z_log_sigma_C_obs", A_n),
  
  # random-effect SDs (needed for SD diagnostics)
  vnames_vec("sigma_A", 4),
  vnames_vec("sigma_B", B_n),
  vnames_vec("sigma_S", S_n),
  vnames_vec("sigma_F", F_n),

  # latent states (length N) used by downstream plots
  vnames_vec("mu_C",    N),
  vnames_vec("C_true",  N),
  vnames_vec("mu_V",    N),
  vnames_vec("V_true",  N),
  vnames_vec("mu_res",  N),
  vnames_vec("RES_star",N),

  # v1p8 residue crop grouping (PrevCr)
  vnames_vec("gamma_PCr_res", PCr_n)
)

vars_scalars <- c(
  # volume / other slopes
  "a_V", "b_V", "beta_vol", "beta_vin", "beta_res_V",
  "mu_beta_irr", "sigma_V", "sigma_VOL_obs",

  # irrigation hierarchy (vector sigma_beta_irr + Cholesky L_beta_irr)
  # sigma_beta_irr is vector length A_n; L_beta_irr is matrix A_n x A_n (Cholesky)
  # (these are used in some diagnostics/plots; harmless to include)
  vnames_vec("sigma_beta_irr", A_n),
  vnames_mat("L_beta_irr", A_n, A_n),

  # crop SDs
  vnames_vec("sigma_Cr", A_n),
  "sigma_Cr_V",

  # GP hyperparams
  "etasq_year", "rhosq_year",

  # concentration SD hyperparams
  "sigma_analyte", "mu_log_sigma_C_obs", "tau_log_sigma_C_obs",

  # residue submodel scalars
  "res_base", "b_res_stir", "sigma_PCr_res", "sigma_res_obs"
)

# Flatten (vars_scalars contains indexed names too)
vars_to_pull <- c(vars_indexed, vars_scalars)

# ---- Pull draws (NO fit$summary) ----
# If you ever want to reduce memory: set max_draws <- 500 and subsample posterior draws
draws_mat <- fit$draws(variables = vars_to_pull, format = "matrix")
n_draws <- nrow(draws_mat)

# ---- Utilities to retrieve indexed/scalar params from draws_mat ----
get_param_mat <- function(draws_mat, par) {
  sel <- grepl(paste0("^", par, "\\["), colnames(draws_mat))
  if (!any(sel)) stop("No indexed columns found for parameter '", par, "'.")
  draws_mat[, sel, drop = FALSE]
}
get_param_vec <- function(draws_mat, par) {
  if (!(par %in% colnames(draws_mat))) stop("No scalar column found for parameter '", par, "'.")
  draws_mat[, par]
}

# ---- analyte-level matrices (draws × A_n)
alpha_mat      <- get_param_mat(draws_mat, "alpha")
beta_stir_mat  <- get_param_mat(draws_mat, "beta_stir")
beta_cin_mat   <- get_param_mat(draws_mat, "beta_cin")
beta_dup_mat   <- get_param_mat(draws_mat, "beta_dup")
beta_res_C_mat <- get_param_mat(draws_mat, "beta_res_C")

# ---- analyte × block random effects: draws × A_n × B_n
gamma_B_mat <- get_param_mat(draws_mat, "gamma_B")
gamma_B_arr <- array(NA_real_, dim = c(n_draws, A_n, B_n))
k <- 1L
for (b in seq_len(B_n)) for (a in seq_len(A_n)) { gamma_B_arr[, a, b] <- gamma_B_mat[, k]; k <- k + 1L }

# ---- analyte × sampler random effects: draws × A_n × S_n
gamma_S_mat <- get_param_mat(draws_mat, "gamma_S")
gamma_S_arr <- array(NA_real_, dim = c(n_draws, A_n, S_n))
k <- 1L
for (s in seq_len(S_n)) for (a in seq_len(A_n)) { gamma_S_arr[, a, s] <- gamma_S_mat[, k]; k <- k + 1L }

# ---- analyte × flume random effects: draws × A_n × F_n
gamma_F_mat <- get_param_mat(draws_mat, "gamma_F")
gamma_F_arr <- array(NA_real_, dim = c(n_draws, A_n, F_n))
k <- 1L
for (f in seq_len(F_n)) for (a in seq_len(A_n)) { gamma_F_arr[, a, f] <- gamma_F_mat[, k]; k <- k + 1L }

# ---- GP year effects: draws × Y_n × A_n
F_mat <- get_param_mat(draws_mat, "F_year")
F_array <- array(NA_real_, dim = c(n_draws, Y_n, A_n))
k <- 1L
for (a in seq_len(A_n)) for (y in seq_len(Y_n)) { F_array[, y, a] <- F_mat[, k]; k <- k + 1L }

# ---- crop effects on concentration: gamma_Cr[A, Cr]
gamma_Cr_mat <- get_param_mat(draws_mat, "gamma_Cr")
if (ncol(gamma_Cr_mat) != A_n * Cr_n) {
  stop("gamma_Cr dimension mismatch: expected ", A_n * Cr_n, " columns but found ", ncol(gamma_Cr_mat), ".")
}
gamma_Cr_arr <- array(NA_real_, dim = c(n_draws, A_n, Cr_n))
k <- 1L
for (cr in seq_len(Cr_n)) for (a in seq_len(A_n)) { gamma_Cr_arr[, a, cr] <- gamma_Cr_mat[, k]; k <- k + 1L }

# ---- crop effects on volume: gamma_Cr_V[Cr]
gamma_Cr_V_mat <- get_param_mat(draws_mat, "gamma_Cr_V")
if (ncol(gamma_Cr_V_mat) != Cr_n) {
  stop("gamma_Cr_V dimension mismatch: expected ", Cr_n, " columns but found ", ncol(gamma_Cr_V_mat), ".")
}

# ---- volume + irrigation terms
beta_vol_vec      <- get_param_vec(draws_mat, "beta_vol")
beta_vin_vec      <- get_param_vec(draws_mat, "beta_vin")
beta_irr_mat      <- get_param_mat(draws_mat, "beta_irr")
a_V_vec           <- get_param_vec(draws_mat, "a_V")
b_V_vec           <- get_param_vec(draws_mat, "b_V")
sigma_V_vec       <- get_param_vec(draws_mat, "sigma_V")
sigma_VOL_obs_vec <- get_param_vec(draws_mat, "sigma_VOL_obs")
beta_res_V_vec    <- get_param_vec(draws_mat, "beta_res_V")
mu_beta_irr_vec   <- get_param_vec(draws_mat, "mu_beta_irr")

# ---- sigma objects (v1p8)
sigma_analyte_vec        <- get_param_vec(draws_mat, "sigma_analyte")
sigma_C_obs_mat          <- get_param_mat(draws_mat, "sigma_C_obs")
mu_log_sigma_C_obs_vec   <- get_param_vec(draws_mat, "mu_log_sigma_C_obs")
tau_log_sigma_C_obs_vec  <- get_param_vec(draws_mat, "tau_log_sigma_C_obs")
z_log_sigma_C_obs_mat    <- get_param_mat(draws_mat, "z_log_sigma_C_obs")
sigma_Cr_mat             <- get_param_mat(draws_mat, "sigma_Cr")
sigma_Cr_V_vec           <- get_param_vec(draws_mat, "sigma_Cr_V")
etasq_year_vec           <- get_param_vec(draws_mat, "etasq_year")
rhosq_year_vec           <- get_param_vec(draws_mat, "rhosq_year")

# ---- latent states used downstream
mu_C_mat    <- get_param_mat(draws_mat, "mu_C")
C_true_mat  <- get_param_mat(draws_mat, "C_true")
mu_V_mat    <- get_param_mat(draws_mat, "mu_V")
V_true_mat  <- get_param_mat(draws_mat, "V_true")
mu_res_mat  <- get_param_mat(draws_mat, "mu_res")
RES_star_mat<- get_param_mat(draws_mat, "RES_star")

# ---- residue submodel (v1p8: previous crop)
res_base_vec      <- get_param_vec(draws_mat, "res_base")
b_res_stir_vec    <- get_param_vec(draws_mat, "b_res_stir")
gamma_PCr_res_mat <- get_param_mat(draws_mat, "gamma_PCr_res")  # draws × PCr_n
sigma_PCr_res_vec <- get_param_vec(draws_mat, "sigma_PCr_res")
sigma_res_obs_vec <- get_param_vec(draws_mat, "sigma_res_obs")

# Alias for downstream chunks that expect "Cr"-named residue group effects
gamma_Cr_res_mat <- gamma_PCr_res_mat
sigma_Cr_res_vec <- sigma_PCr_res_vec

# ---- Assemble posterior list (keep names used by v1p7p5 downstream code)
post <- list(
  alpha         = alpha_mat,
  beta_stir     = beta_stir_mat,
  beta_cin      = beta_cin_mat,
  beta_dup      = beta_dup_mat,
  beta_res_C    = beta_res_C_mat,

  gamma_B       = gamma_B_arr,
  gamma_S       = gamma_S_arr,
  gamma_F       = gamma_F_arr,

  gamma_Cr      = gamma_Cr_arr,       # draws × A_n × Cr_n
  gamma_Cr_V    = gamma_Cr_V_mat,     # draws × Cr_n

  beta_vol      = beta_vol_vec,
  beta_vin      = beta_vin_vec,
  beta_irr      = beta_irr_mat,
  mu_beta_irr   = mu_beta_irr_vec,
  a_V           = a_V_vec,
  b_V           = b_V_vec,
  sigma_V       = sigma_V_vec,
  sigma_VOL_obs = sigma_VOL_obs_vec,
  beta_res_V    = beta_res_V_vec,

  # C model SDs
  sigma_analyte = sigma_analyte_vec,
  sigma_C_obs   = sigma_C_obs_mat,
  mu_log_sigma_C_obs  = mu_log_sigma_C_obs_vec,
  tau_log_sigma_C_obs = tau_log_sigma_C_obs_vec,
  z_log_sigma_C_obs   = z_log_sigma_C_obs_mat,

  # crop SDs
  sigma_Cr      = sigma_Cr_mat,
  sigma_Cr_V    = sigma_Cr_V_vec,

  # GP
  F_year        = F_array,
  etasq_year    = etasq_year_vec,
  rhosq_year    = rhosq_year_vec,

  # latent states used by plots
  mu_C          = mu_C_mat,
  C_true        = C_true_mat,
  mu_V          = mu_V_mat,
  V_true        = V_true_mat,
  mu_res        = mu_res_mat,
  RES_star      = RES_star_mat,

  # residue submodel (aliased to legacy names)
  res_base      = res_base_vec,
  b_res_stir    = b_res_stir_vec,
  gamma_Cr_res  = gamma_Cr_res_mat,
  sigma_Cr_res  = sigma_Cr_res_vec,
  sigma_res_obs = sigma_res_obs_vec
)

#draws_mat <- pull_draws_mat(vars_to_pull)
n_draws <- nrow(draws_mat)

# ---- Helpers to extract scalars / indexed params ----
get_param_mat <- function(dm, par) {
  sel <- grepl(paste0("^", par, "\\["), colnames(dm))
  if (!any(sel)) stop("No indexed columns found for parameter '", par, "'.")
  dm[, sel, drop = FALSE]
}
get_param_vec <- function(dm, par) {
  if (!(par %in% colnames(dm))) stop("No scalar column found for parameter '", par, "'.")
  dm[, par]
}

# ---- Analyte-level (draws × A_n) ----
alpha_mat      <- get_param_mat(draws_mat, "alpha")
beta_stir_mat  <- get_param_mat(draws_mat, "beta_stir")
beta_cin_mat   <- get_param_mat(draws_mat, "beta_cin")
beta_dup_mat   <- get_param_mat(draws_mat, "beta_dup")
beta_res_C_mat <- get_param_mat(draws_mat, "beta_res_C")

# ---- Random effects: gamma_B[A,B], gamma_S[A,S], gamma_F[A,F] ----
gamma_B_mat <- get_param_mat(draws_mat, "gamma_B")
gamma_B_arr <- array(NA_real_, dim = c(n_draws, A_n, B_n))
k <- 1L
for (b in seq_len(B_n)) for (a in seq_len(A_n)) { gamma_B_arr[, a, b] <- gamma_B_mat[, k]; k <- k + 1L }

gamma_S_mat <- get_param_mat(draws_mat, "gamma_S")
gamma_S_arr <- array(NA_real_, dim = c(n_draws, A_n, S_n))
k <- 1L
for (s in seq_len(S_n)) for (a in seq_len(A_n)) { gamma_S_arr[, a, s] <- gamma_S_mat[, k]; k <- k + 1L }

gamma_F_mat <- get_param_mat(draws_mat, "gamma_F")
gamma_F_arr <- array(NA_real_, dim = c(n_draws, A_n, F_n))
k <- 1L
for (f in seq_len(F_n)) for (a in seq_len(A_n)) { gamma_F_arr[, a, f] <- gamma_F_mat[, k]; k <- k + 1L }

# ---- GP year effects: F_year[Y,A] ----
F_mat <- get_param_mat(draws_mat, "F_year")
F_array <- array(NA_real_, dim = c(n_draws, Y_n, A_n))
k <- 1L
for (a in seq_len(A_n)) for (y in seq_len(Y_n)) { F_array[, y, a] <- F_mat[, k]; k <- k + 1L }

# ---- Crop effects on concentration: gamma_Cr[A,Cr] ----
gamma_Cr_mat <- get_param_mat(draws_mat, "gamma_Cr")
if (ncol(gamma_Cr_mat) != A_n * Cr_n) {
  stop("gamma_Cr dimension mismatch: expected ", A_n * Cr_n, " but found ", ncol(gamma_Cr_mat), ".")
}
gamma_Cr_arr <- array(NA_real_, dim = c(n_draws, A_n, Cr_n))
k <- 1L
for (cr in seq_len(Cr_n)) for (a in seq_len(A_n)) { gamma_Cr_arr[, a, cr] <- gamma_Cr_mat[, k]; k <- k + 1L }

# ---- Crop effects on volume: gamma_Cr_V[Cr] ----
gamma_Cr_V_mat <- get_param_mat(draws_mat, "gamma_Cr_V")
if (ncol(gamma_Cr_V_mat) != Cr_n) {
  stop("gamma_Cr_V dimension mismatch: expected ", Cr_n, " but found ", ncol(gamma_Cr_V_mat), ".")
}

# ---- Irrigation hierarchy: beta_irr[A], plus mu/sigma/L ----
beta_irr_mat <- get_param_mat(draws_mat, "beta_irr")
if (ncol(beta_irr_mat) != A_n) {
  stop("beta_irr dimension mismatch: expected ", A_n, " but found ", ncol(beta_irr_mat), ".")
}

mu_beta_irr_vec   <- get_param_vec(draws_mat, "mu_beta_irr")
sigma_beta_irr_mat <- get_param_mat(draws_mat, "sigma_beta_irr")
if (ncol(sigma_beta_irr_mat) != A_n) {
  stop("sigma_beta_irr dimension mismatch: expected ", A_n, " but found ", ncol(sigma_beta_irr_mat), ".")
}

L_beta_irr_mat <- get_param_mat(draws_mat, "L_beta_irr")
if (ncol(L_beta_irr_mat) != A_n * A_n) {
  stop("L_beta_irr dimension mismatch: expected ", A_n*A_n, " but found ", ncol(L_beta_irr_mat), ".")
}
L_beta_irr_arr <- array(NA_real_, dim = c(n_draws, A_n, A_n))
k <- 1L
for (j in seq_len(A_n)) for (i in seq_len(A_n)) { L_beta_irr_arr[, i, j] <- L_beta_irr_mat[, k]; k <- k + 1L }

# ---- GP correlation + SD ----
sigma_Agp_mat <- get_param_mat(draws_mat, "sigma_Agp")
if (ncol(sigma_Agp_mat) != A_n) stop("sigma_Agp dimension mismatch: expected ", A_n, " but found ", ncol(sigma_Agp_mat), ".")
L_corr_Agp_mat <- get_param_mat(draws_mat, "L_corr_Agp")
if (ncol(L_corr_Agp_mat) != A_n * A_n) stop("L_corr_Agp dimension mismatch: expected ", A_n*A_n, " but found ", ncol(L_corr_Agp_mat), ".")
L_corr_Agp_arr <- array(NA_real_, dim = c(n_draws, A_n, A_n))
k <- 1L
for (j in seq_len(A_n)) for (i in seq_len(A_n)) { L_corr_Agp_arr[, i, j] <- L_corr_Agp_mat[, k]; k <- k + 1L }

# ---- Scalars + SD structures ----
beta_vol_vec       <- get_param_vec(draws_mat, "beta_vol")
beta_vin_vec       <- get_param_vec(draws_mat, "beta_vin")
a_V_vec            <- get_param_vec(draws_mat, "a_V")
b_V_vec            <- get_param_vec(draws_mat, "b_V")
sigma_V_vec        <- get_param_vec(draws_mat, "sigma_V")
sigma_VOL_obs_vec  <- get_param_vec(draws_mat, "sigma_VOL_obs")
beta_res_V_vec     <- get_param_vec(draws_mat, "beta_res_V")

sigma_A_mat <- get_param_mat(draws_mat, "sigma_A")
sigma_B_mat <- get_param_mat(draws_mat, "sigma_B")
sigma_S_mat <- get_param_mat(draws_mat, "sigma_S")
sigma_F_mat <- get_param_mat(draws_mat, "sigma_F")

sigma_Cr_mat <- get_param_mat(draws_mat, "sigma_Cr")
sigma_Cr_V_vec <- get_param_vec(draws_mat, "sigma_Cr_V")

etasq_year_vec <- get_param_vec(draws_mat, "etasq_year")
rhosq_year_vec <- get_param_vec(draws_mat, "rhosq_year")

sigma_analyte_vec <- get_param_vec(draws_mat, "sigma_analyte")
sigma_C_obs_mat   <- get_param_mat(draws_mat, "sigma_C_obs")
mu_log_sigma_C_obs_vec  <- get_param_vec(draws_mat, "mu_log_sigma_C_obs")
tau_log_sigma_C_obs_vec <- get_param_vec(draws_mat, "tau_log_sigma_C_obs")
z_log_sigma_C_obs_mat   <- get_param_mat(draws_mat, "z_log_sigma_C_obs")

# ---- Latent states (draws × N) ----
mu_C_mat   <- get_param_mat(draws_mat, "mu_C")
C_true_mat <- get_param_mat(draws_mat, "C_true")
mu_V_mat   <- get_param_mat(draws_mat, "mu_V")
V_true_mat <- get_param_mat(draws_mat, "V_true")
mu_res_mat <- get_param_mat(draws_mat, "mu_res")
RES_star_mat <- get_param_mat(draws_mat, "RES_star")

# ---- Residue submodel (v1p8: previous crop) ----
res_base_vec   <- get_param_vec(draws_mat, "res_base")
b_res_stir_vec <- get_param_vec(draws_mat, "b_res_stir")
gamma_PCr_res_mat <- get_param_mat(draws_mat, "gamma_PCr_res")
if (ncol(gamma_PCr_res_mat) != PCr_n) {
  stop("gamma_PCr_res dimension mismatch: expected ", PCr_n, " but found ", ncol(gamma_PCr_res_mat), ".")
}
sigma_PCr_res_vec <- get_param_vec(draws_mat, "sigma_PCr_res")
sigma_res_obs_vec <- get_param_vec(draws_mat, "sigma_res_obs")

# ---- Assemble posterior list (keep v1p7p5 names where downstream code expects them) ----
post <- list(
  # concentration fixed effects
  alpha      = alpha_mat,
  beta_stir  = beta_stir_mat,
  beta_cin   = beta_cin_mat,
  beta_dup   = beta_dup_mat,
  beta_res_C = beta_res_C_mat,

  # random effects
  gamma_B = gamma_B_arr,
  gamma_S = gamma_S_arr,
  gamma_F = gamma_F_arr,

  # crop effects
  gamma_Cr   = gamma_Cr_arr,     # draws × A × Cr
  gamma_Cr_V = gamma_Cr_V_mat,   # draws × Cr

  # volume / irrigation
  beta_vol      = beta_vol_vec,
  beta_vin      = beta_vin_vec,
  beta_irr      = beta_irr_mat,
  mu_beta_irr   = mu_beta_irr_vec,
  sigma_beta_irr = sigma_beta_irr_mat,
  L_beta_irr    = L_beta_irr_arr,
  a_V           = a_V_vec,
  b_V           = b_V_vec,
  sigma_V       = sigma_V_vec,
  sigma_VOL_obs = sigma_VOL_obs_vec,
  beta_res_V    = beta_res_V_vec,

  # SDs for group effects
  sigma_A = sigma_A_mat,
  sigma_B = sigma_B_mat,
  sigma_S = sigma_S_mat,
  sigma_F = sigma_F_mat,

  # concentration SD structure
  sigma_analyte = sigma_analyte_vec,
  sigma_C_obs   = sigma_C_obs_mat,
  mu_log_sigma_C_obs  = mu_log_sigma_C_obs_vec,
  tau_log_sigma_C_obs = tau_log_sigma_C_obs_vec,
  z_log_sigma_C_obs   = z_log_sigma_C_obs_mat,

  # crop SDs
  sigma_Cr   = sigma_Cr_mat,
  sigma_Cr_V = sigma_Cr_V_vec,

  # GP
  F_year      = F_array,
  etasq_year  = etasq_year_vec,
  rhosq_year  = rhosq_year_vec,
  sigma_Agp   = sigma_Agp_mat,
  L_corr_Agp  = L_corr_Agp_arr,

  # latent states
  mu_C    = mu_C_mat,
  C_true  = C_true_mat,
  mu_V    = mu_V_mat,
  V_true  = V_true_mat,
  mu_res  = mu_res_mat,
  RES_star = RES_star_mat,

  # residue submodel (v1p8-native names)
  res_base      = res_base_vec,
  b_res_stir    = b_res_stir_vec,
  gamma_PCr_res = gamma_PCr_res_mat,
  sigma_PCr_res = sigma_PCr_res_vec,
  sigma_res_obs = sigma_res_obs_vec,

  # backwards-compatible aliases (so older plotting chunks still work)
  gamma_Cr_res  = gamma_PCr_res_mat,
  sigma_Cr_res  = sigma_PCr_res_vec,

  # residue placeholder flag for downstream summaries
  res_is_placeholder = res_is_placeholder
)

# Quick sanity print (fast, no summary())
cat("\n[v1p8] post assembled.\n")
cat("  draws:", n_draws, " A_n:", A_n, " Y_n:", Y_n, " Cr_n:", Cr_n, " PrevCr_n:", PCr_n, "\n")
cat("  residue placeholder:", post$res_is_placeholder, "\n")
```


## 5a. Posterior effects of residue and crop

The v1p8 model includes (i) an analyte-specific residue slope on concentration (`beta_res_C[A]`), (ii) a scalar residue slope on volume (`beta_res_V`), (iii) analyte-specific crop intercepts on concentration (`gamma_Cr[A,Cr]`), and (iv) crop intercepts on volume (`gamma_Cr_V[Cr]`). The plots below provide quick, publication-ready summaries of these terms.

```{r residue effects on concentration}
# Residue effect on concentration by analyte
# ---- Prepare posterior draws ----
post_resC_df <- as.data.frame(post$beta_res_C)
colnames(post_resC_df) <- as.character(analyte_lookup$analyte_abbr)

post_resC_long <- post_resC_df %>%
  pivot_longer(
    cols = everything(),
    names_to = "analyte_abbr",
    values_to = "beta_resC"
  )

# ---- Reorder analytes by posterior median (base R, no forcats) ----
analyte_order <- post_resC_long %>%
  group_by(analyte_abbr) %>%
  summarise(med = median(beta_resC)) %>%
  arrange(med) %>%
  pull(analyte_abbr)

post_resC_long$analyte_abbr <- factor(
  post_resC_long$analyte_abbr,
  levels = analyte_order
)

# ---- Ridgeline plot ----
p_resC_ridge <- ggplot(post_resC_long, aes(x = beta_resC, y = analyte_abbr)) +
  geom_vline(xintercept = 0, linetype = 2, linewidth=1) +
  stat_density_ridges(
    fill = "darkred",
    alpha = 0.55,
    scale = 1.1,
    rel_min_height = 0.01,
    quantile_lines = TRUE,
    quantiles = 2,
    quantile_color = "black",
    quantile_linetype = 2,
    quantile_linewidth = 1,
    bandwidth = 0.02
  ) +
  labs(
    x = "Residue effect on concentration",
    y = "Analyte",
    #title = "Posterior residue effects on concentration (v1p8)",
    subtitle = if (res_is_placeholder)
      "Residue is currently a placeholder (all zeros), effects are not identifiable"
      else NULL
  ) +
  theme_ridges(grid = TRUE, center_axis_labels = TRUE) +
  theme(
    legend.position = "none",
    text = element_text(size = 14)
  )

p_resC_ridge

ggsave(
  filename = "../figs/residue_effect_on_concentration_by_analyte_v1p8.jpg",
  plot = p_resC_ridge,
  width = 10,
  height = 7,
  dpi = 300
)

```

```{r residue effect on volume}
# Residue effect on volume (scalar slope) posterior density
# If placeholder, report and plot a spike at zero.

b_resV <- if (isTRUE(res_is_placeholder)) rep(0, 200) else post$beta_res_V

dens <- stats::density(b_resV)

plot(
  dens,
  type = "l",
  lwd  = 3,
  col  = "darkred",
  xlab = "Residue effect on Volume",
  ylab = "Density",
  main = "Posterior residue effect on volume",
  bty  = "l",
  # ensure xlim always shows zero
  xlim = range(c(dens$x, 0))
)
abline(v = 0, lty = 2)

jpeg("../figs/residue_effect_on_volume_density_v1p8.jpeg", width = 2000, height = 1300, res = 200)
plot(
  dens,
  type = "l",
  lwd  = 3,
  col  = "darkred",
  xlab = "Residue effect on Volume",
  ylab = "Density",
  main = "Posterior residue effect on volume",
  bty  = "l",
  # ensure xlim always shows zero
  xlim = range(c(dens$x, 0))
)
abline(v = 0, lty = 2)
dev.off()
```

```{r crop effects on concentration}
# ------------------------------------------------------------
# Crop effects on concentration: posterior ridgelines by crop
# post$gamma_Cr is draws × A_n × Cr_n
# ------------------------------------------------------------

# Build a long draw table: one row per (draw, analyte, crop)
crop_levels <- as.character(crop_lookup$Crop)
analyte_levels <- as.character(analyte_lookup$analyte_abbr)

cropC_long <- bind_rows(lapply(seq_len(Cr_n), function(cr) {
  m <- post$gamma_Cr[, , cr, drop = FALSE]      # draws × A × 1
  m2 <- m[, , 1]                               # draws × A
  df <- as.data.frame(m2)
  colnames(df) <- analyte_levels
  df %>%
    pivot_longer(
      cols = everything(),
      names_to = "analyte_abbr",
      values_to = "crop_eff"
    ) %>%
    mutate(Crop = crop_levels[cr])
}))

# Make Crop a factor with your desired ordering
cropC_long$Crop <- factor(cropC_long$Crop, levels = crop_levels)

# Choose ONE analyte order used in every facet (recommended for cross-panel comparison):
analyte_order_global <- cropC_long %>%
  group_by(analyte_abbr) %>%
  summarise(med = median(crop_eff), .groups = "drop") %>%
  arrange(med) %>%
  pull(analyte_abbr)

cropC_long$analyte_abbr <- factor(cropC_long$analyte_abbr, levels = analyte_order_global)

# ------------------------------------------------------------
# NEW: pick a tighter x-axis range from posterior quantiles
# - uses pooled draws across all analytes/crops
# - keeps 0 inside the view
# ------------------------------------------------------------
q_lo <- unname(quantile(cropC_long$crop_eff, probs = 0.005, na.rm = TRUE))
q_hi <- unname(quantile(cropC_long$crop_eff, probs = 0.995, na.rm = TRUE))

# symmetric range around 0, with a small pad so ridges aren't clipped
half_range <- max(abs(c(q_lo, q_hi)), 0.05)   # 0.05 is a safety floor
pad <- 0.05 * half_range
x_lim <- c(-half_range - pad, half_range + pad)

# ---- Faceted ridgeline plot ----
p_cropC_ridge <- ggplot(cropC_long, aes(x = crop_eff, y = analyte_abbr)) +
  geom_vline(xintercept = 0, linetype = 2, linewidth = 0.8) +
  stat_density_ridges(
    fill = "darkred",
    alpha = 0.55,
    scale = 1.05,
    rel_min_height = 0.01,
    quantile_lines = TRUE,
    quantiles = 2,
    quantile_color = "black",
    quantile_linetype = 2,
    quantile_linewidth = 0.9,
    bandwidth = 0.02
  ) +
  facet_wrap(~ Crop, ncol = 3, scales = "fixed") +
  coord_cartesian(xlim = x_lim) +
  labs(
    x = "Crop effect on concentration",
    y = "Analyte",
    subtitle = "Posterior distributions of analyte-specific crop intercepts"
  ) +
  theme_ridges(grid = TRUE, center_axis_labels = TRUE) +
  theme(
    legend.position = "none",
    text = element_text(size = 12),
    strip.text = element_text(size = 12),
    axis.text.y = element_text(size = 9)
  )

p_cropC_ridge

ggsave(
  filename = "../figs/crop_effects_on_concentration_by_analyte_v1p8.jpg",
  plot = p_cropC_ridge,
  width = 9,
  height = 12,
  dpi = 300
)

```

```{r crop effects on volume}
# ------------------------------------------------------------
# Crop effects on volume: ridgeline posterior by crop
# post$gamma_Cr_V is draws × Cr_n
# ------------------------------------------------------------

# Long posterior draw table
post_cropV_df <- as.data.frame(post$gamma_Cr_V)
colnames(post_cropV_df) <- as.character(crop_lookup$Crop)

post_cropV_long <- post_cropV_df %>%
  pivot_longer(
    cols = everything(),
    names_to = "Crop",
    values_to = "crop_eff_V"
  )

# Reorder crops by posterior median (base R, no forcats)
crop_order <- post_cropV_long %>%
  group_by(Crop) %>%
  summarise(med = median(crop_eff_V), .groups = "drop") %>%
  arrange(med) %>%
  pull(Crop)

post_cropV_long$Crop <- factor(post_cropV_long$Crop, levels = crop_order)

# Ridgeline plot
p_cropV_ridge <- ggplot(post_cropV_long, aes(x = crop_eff_V, y = Crop)) +
  geom_vline(xintercept = 0, linetype = 2, linewidth = 1) +
  stat_density_ridges(
    fill = "darkred",
    alpha = 0.55,
    scale = 1.1,
    rel_min_height = 0.01,
    quantile_lines = TRUE,
    quantiles = 2,
    quantile_color = "black",
    quantile_linetype = 2,
    quantile_linewidth = 1,
    bandwidth = 0.02
  ) +
  labs(
    x = "Crop effect on volume",
    y = "Crop",
    #title = "Posterior crop effects on volume"
  ) +
  theme_ridges(grid = TRUE, center_axis_labels = TRUE) +
  theme(
    legend.position = "none",
    text = element_text(size = 14)
  )

p_cropV_ridge

ggsave(
  filename = "../figs/crop_effects_on_volume_v1p8.jpg",
  plot = p_cropV_ridge,
  width = 10,
  height = 6,
  dpi = 300
)

```



---



# 4.5 Plot posterior slopes for irrigation-count effects on concentration

```{r irrigation effects on conc}
# ------------------------------------------------------------
# Irrigation-count effects on concentration by analyte
# beta_irr_mat is draws × A_n (MVN prior across analytes in v1p8)
# Note: Stan standardizes IRR internally (IRR_z), so beta_irr is per +1 SD(IRR)
# ------------------------------------------------------------

# Long posterior draw table
post_irr_df <- as.data.frame(beta_irr_mat)
colnames(post_irr_df) <- as.character(analytes_keep)

post_irr_long <- post_irr_df %>%
  tidyr::pivot_longer(
    cols = everything(),
    names_to = "Analyte",
    values_to = "beta_irr"
  )

# Reorder analytes by posterior median (base R, no forcats)
analyte_order <- post_irr_long %>%
  dplyr::group_by(Analyte) %>%
  dplyr::summarise(med = median(beta_irr), .groups = "drop") %>%
  dplyr::arrange(med) %>%
  dplyr::pull(Analyte)

post_irr_long$Analyte <- factor(post_irr_long$Analyte, levels = analyte_order)

# Ridgeline plot (match crop style)
p_irr_ridge <- ggplot(post_irr_long, aes(x = beta_irr, y = Analyte)) +
  geom_vline(xintercept = 0, linetype = 2, linewidth = 1) +
  ggridges::stat_density_ridges(
    fill = "darkred",
    alpha = 0.55,
    scale = 1.1,
    rel_min_height = 0.01,
    quantile_lines = TRUE,
    quantiles = 2,
    quantile_color = "black",
    quantile_linetype = 2,
    quantile_linewidth = 1,
    bandwidth = 0.02
  ) +
  labs(
    x = "Effect of Irrigation Number on Concentration",
    y = "Analyte"
    # title = "Posterior irrigation-number effects on concentration"
  ) +
  theme_ridges(grid = TRUE, center_axis_labels = TRUE) +
  theme(
    legend.position = "none",
    text = element_text(size = 14)
  )

p_irr_ridge

ggsave(
  filename = "../figs/irr_effects_on_concentration_v1p8.jpg",
  plot = p_irr_ridge,
  width = 10,
  height = 6,
  dpi = 300
)

```


# 5. Plot posterior slopes for STIR effects on concentration

```{r stir effects on conc}
# ------------------------------------------------------------
# STIR effects on concentration by analyte: posterior ridgelines
# Uses post$beta_stir (draws × A_n)
# Overwrites: ../figs/load1p8_STIReffects.jpeg
# ------------------------------------------------------------

# ---- Prepare posterior draws ----
post_stir_df <- as.data.frame(post$beta_stir)
colnames(post_stir_df) <- as.character(analyte_lookup$analyte_abbr)

post_stir_long <- post_stir_df %>%
  pivot_longer(
    cols = everything(),
    names_to = "analyte_abbr",
    values_to = "beta_stir"
  )

# ---- Order analytes by posterior mean (base R, no forcats) ----
analyte_order <- post_stir_long %>%
  group_by(analyte_abbr) %>%
  summarise(mu = mean(beta_stir), .groups = "drop") %>%
  arrange(mu) %>%
  pull(analyte_abbr)

post_stir_long$analyte_abbr <- factor(
  post_stir_long$analyte_abbr,
  levels = analyte_order
)

# ---- Ridgeline plot ----
p_stir_ridge <- ggplot(post_stir_long, aes(x = beta_stir, y = analyte_abbr)) +
  geom_vline(xintercept = 0, linetype = 2, linewidth = 1) +
  stat_density_ridges(
    fill = "darkred",
    alpha = 0.55,
    scale = 1.1,
    rel_min_height = 0.01,
    quantile_lines = TRUE,
    quantiles = 2,
    quantile_color = "black",
    quantile_linetype = 2,
    quantile_linewidth = 1,
    bandwidth = 0.01
  ) +
  labs(
    x = "Effect of STIR on Concentration",
    y = "Analyte (ordered by mean effect)",
    title = "Analyte-specific STIR effects with posterior medians"
  ) +
  theme_ridges(grid = TRUE, center_axis_labels = TRUE) +
  theme(
    legend.position = "none",
    text = element_text(size = 14),
    axis.text.y = element_text(size = 10)
  )

p_stir_ridge

# ---- Save (overwrite existing) ----
ggsave(
  filename = "../figs/load1p8_STIReffects.jpeg",
  plot = p_stir_ridge,
  width = 9,
  height = 9,
  dpi = 300
)

```

# ------------------------------------------------------------
# STIR effects on volume
# ------------------------------------------------------------

```{r stir effects on volume}
plot_stir_volume_effect <- function(fit, param = "b_V",
                                    file = NULL,
                                    main = "Posterior effect of STIR on runoff volume",
                                    xlab = "Effect of STIR on runoff volume",
                                    ylab = "Density") {

  if (inherits(fit, "CmdStanMCMC")) {
    draws <- fit$draws(variables = param, format = "matrix")
    if (!(param %in% colnames(draws))) {
      stop("Parameter '", param, "' not found in cmdstan draws.")
    }
    b <- as.numeric(draws[, param])
  } else if (inherits(fit, "stanfit")) {
    s <- rstan::extract(fit, pars = param, permuted = TRUE)
    if (!(param %in% names(s))) stop("Parameter '", param, "' not found in stanfit.")
    b <- as.numeric(s[[param]])
  } else if (is.numeric(fit)) {
    b <- as.numeric(fit)
  } else {
    stop("Unsupported object type for 'fit'. Use a CmdStanMCMC, stanfit, or numeric vector.")
  }

  b <- b[is.finite(b)]

  if (!is.null(file)) {
    jpeg(file, width = 2000, height = 1300, res = 200)
    on.exit(dev.off(), add = TRUE)
  }

  dens <- stats::density(b)

  plot(
    dens,
    type = "l",
    lwd  = 4,
    col  = "darkred",
    xlab = xlab,
    ylab = ylab,
    main = main,
    bty  = "l",
    xlim = range(c(dens$x, 0))
  )

  abline(v = 0, lty = 2)

  invisible(b)
}

# Plot to device
plot_stir_volume_effect(fit, param = "b_V")

# Save to file (v1p8 filename)
plot_stir_volume_effect(
  fit,
  param = "b_V",
  file = "../figs/v1p8_post_STIR_effect_on_volume.jpg"
)
```

# ------------------------------------------------------------
# STIR effects on residue
# ------------------------------------------------------------

```{r}
# STIR effect on residue (logit-normal submodel)
# Scalar slope: b_res_stir

b_res_stir <- post$b_res_stir

dens <- stats::density(b_res_stir)

# ---- Plot to Rmd output ----
plot(
  dens,
  type = "l",
  lwd  = 3,
  col  = "darkred",
  xlab = "STIR effect on logit(Residue proportion)",
  ylab = "Density",
  main = "Posterior STIR effect on residue",
  bty  = "l",
  # ensure zero is always visible
  xlim = range(c(dens$x, 0))
)

abline(v = 0, lty = 2)

# ---- Save identical plot to file ----
jpeg("../figs/stir_effect_on_residue_density_v1p8.jpeg",
     width = 2000, height = 1300, res = 200)

plot(
  dens,
  type = "l",
  lwd  = 3,
  col  = "darkred",
  xlab = "STIR effect on logit(Residue proportion)",
  ylab = "Density",
  main = "Posterior STIR effect on residue",
  bty  = "l",
  xlim = range(c(dens$x, 0))
)

abline(v = 0, lty = 2)

dev.off()
```


## 6. Generative STIR → Load curves by analyte --------------------------------

```{r load-generative-curves}
## Scenario-based STIR -> Load link + posterior predictive curves ----------
## v1p8 (dual ribbons):
##   1) "mean" ribbon = uncertainty in expected mean relationship (link-style)
##   2) "pp" ribbon   = posterior predictive including latent process + obs noise (sim-style)
##
## Notes:
## - Uses sigma_V + sigma_VOL_obs for volume (latent + obs)
## - Uses sigma_analyte + sigma_C_obs[A] for concentration (latent + obs)
## - Residue submodel is NOT used; RES is treated as an input covariate.
## - v1p8: irrigation effect is analyte-specific and IRR is standardized (IRR_z).

# 1. Back-transform stats (only computed once)
c_stats <- d_mod %>%
  dplyr::group_by(A) %>%
  dplyr::summarize(
    c_mean = mean(Result_mg_L, na.rm = TRUE),
    c_sd   = sd(Result_mg_L,   na.rm = TRUE),
    .groups = "drop"
  )

V_mean <- mean(d_mod$Volume, na.rm = TRUE)
V_sd   <- sd(d_mod$Volume, na.rm = TRUE)

# STIR raw mean/sd for Season_STIR_toDate
stir_mean <- mean(d_mod$Season_STIR_toDate, na.rm = TRUE)
stir_sd   <- sd(d_mod$Season_STIR_toDate, na.rm = TRUE)

# Empirical means for averaging when left NULL
IRR_bar_raw <- mean(d_mod$irr_num, na.rm = TRUE)
DUP_bar     <- mean(d_mod$dup,     na.rm = TRUE)

# v1p8: IRR standardization to match Stan IRR_z
irr_mean <- mean(d_mod$irr_num, na.rm = TRUE)
irr_sd   <- sd(d_mod$irr_num,   na.rm = TRUE)
if (!is.finite(irr_sd) || irr_sd == 0) irr_sd <- 1
IRR_bar_z <- (IRR_bar_raw - irr_mean) / irr_sd

# Empirical mean residue proportion when res_z is NULL
RES_bar <- if ("RES_input" %in% names(d_mod)) {
  mean(d_mod$RES_input, na.rm = TRUE)
} else if ("RES" %in% names(d_mod)) {
  mean(d_mod$RES, na.rm = TRUE)
} else if ("residue_prop" %in% names(d_mod)) {
  mean(d_mod$residue_prop, na.rm = TRUE)
} else if ("RES_z" %in% names(d_mod)) {
  mean(d_mod$RES_z, na.rm = TRUE)
} else {
  NA_real_
}

if (!is.finite(RES_bar)) stop("Could not compute RES_bar: d_mod needs RES_input, RES, residue_prop, or RES_z as a residue proportion in (0,1).")
if (RES_bar <= 0 || RES_bar >= 1) stop("RES_bar is not in (0,1). Check that residue is stored as a proportion, not percent.")

# Crop weights for "average over crops" scenario
crop_wts <- d_mod %>%
  dplyr::count(Cr, name = "n") %>%
  dplyr::mutate(w = n / sum(n)) %>%
  dplyr::arrange(Cr)

simulate_load_curve_single <- function(
  post,
  analyte,
  STIR_raw_seq = seq(0, 300, length.out = 60),
  block      = NULL,
  sampler    = NULL,
  flume      = NULL,
  irrigation = NULL,
  duplicate  = NULL,
  crop       = NULL,
  res_z      = NULL,   # NULL -> RES_bar; 0 -> 1e-4; (0,1) fixed proportion
  cin_z      = 0,
  year       = NULL,
  n_draws    = 2000,
  ci_prob    = 0.95
) {

  if (!is.numeric(ci_prob) || length(ci_prob) != 1L || ci_prob <= 0 || ci_prob >= 1) {
    stop("ci_prob must be a single numeric value strictly between 0 and 1.")
  }

  # ---- required sigmas in post (fail fast if missing) ----
  need <- c("sigma_V", "sigma_VOL_obs", "sigma_analyte")
  miss <- setdiff(need, names(post))
  if (length(miss) > 0) stop("post is missing required parameter(s): ", paste(miss, collapse = ", "))

  # concentration obs sd is analyte-specific
  if (!("sigma_C_obs" %in% names(post))) {
    stop("post is missing sigma_C_obs (analyte-specific concentration observation SD).")
  }

  # Map analyte label -> index A
  j_idx <- analyte_lookup$A[analyte_lookup$analyte_abbr == analyte]
  if (length(j_idx) != 1L) stop("analyte not found in analyte_lookup")

  # Stats for this analyte
  stats_j  <- dplyr::filter(c_stats, A == j_idx)
  c_mean_j <- stats_j$c_mean
  c_sd_j   <- stats_j$c_sd

  # STIR in z units
  STIR_z_seq <- (STIR_raw_seq - stir_mean) / stir_sd

  # Total draws and subsample index
  n_draws_total <- nrow(post$alpha)
  draw_idx <- if (n_draws_total > n_draws) sample.int(n_draws_total, n_draws) else seq_len(n_draws_total)
  n_s <- length(draw_idx)

  # ---------------- Random effects / covariate settings ----------------

  # Block (Rep)
  if (is.null(block)) {
    gamma_B_draws <- post$gamma_B[draw_idx, j_idx, , drop = FALSE]   # draws × 1 × B_n
    gamma_B_star  <- apply(gamma_B_draws, 1, mean)
    block_label   <- "avg_all_blocks"
  } else {
    B_idx <- block_lookup$B[block_lookup$Rep == block]
    if (length(B_idx) != 1L) stop("block not found in block_lookup")
    gamma_B_star <- post$gamma_B[draw_idx, j_idx, B_idx]
    block_label  <- as.character(block)
  }

  # Sampler method
  if (is.null(sampler)) {
    gamma_S_draws <- post$gamma_S[draw_idx, j_idx, , drop = FALSE]
    gamma_S_star  <- apply(gamma_S_draws, 1, mean)
    sampler_label <- "avg_all_samplers"
  } else {
    S_idx <- sampler_lookup$S[sampler_lookup$SampleMethod == sampler]
    if (length(S_idx) != 1L) stop("sampler not found in sampler_lookup")
    gamma_S_star  <- post$gamma_S[draw_idx, j_idx, S_idx]
    sampler_label <- as.character(sampler)
  }

  # Flume type
  if (is.null(flume)) {
    gamma_F_draws <- post$gamma_F[draw_idx, j_idx, , drop = FALSE]
    gamma_F_star  <- apply(gamma_F_draws, 1, mean)
    flume_label   <- "avg_all_flumes"
  } else {
    Fu_idx <- flume_lookup$Fu[flume_lookup$FlumeMethod == flume]
    if (length(Fu_idx) != 1L) stop("flume not found in flume_lookup")
    gamma_F_star  <- post$gamma_F[draw_idx, j_idx, Fu_idx]
    flume_label   <- as.character(flume)
  }

  # Irrigation (numeric)
  # IMPORTANT: convert raw irrigation count -> IRR_z scale that Stan uses
  if (is.null(irrigation)) {
    IRR_star_raw     <- IRR_bar_raw
    IRR_star_z       <- IRR_bar_z
    irrigation_label <- "avg_all_irrigations"
  } else {
    irr_val <- irrigation_lookup$irr_num[irrigation_lookup$Irrigation == irrigation]
    if (length(irr_val) != 1L) stop("irrigation not found in irrigation_lookup")
    IRR_star_raw     <- irr_val
    IRR_star_z       <- (IRR_star_raw - irr_mean) / irr_sd
    irrigation_label <- as.character(irrigation)
  }

  # Duplicate status (numeric 0/1 or mean)
  if (is.null(duplicate)) {
    DUP_star        <- DUP_bar
    duplicate_label <- "avg_duplicates"
  } else {
    DUP_star        <- as.numeric(duplicate)
    duplicate_label <- ifelse(as.logical(duplicate), "TRUE", "FALSE")
  }

  # Crop: affects BOTH volume and concentration
  if (is.null(crop)) {
    w <- crop_wts$w
    if (length(w) != ncol(post$gamma_Cr_V)) stop("crop weight length != Cr_n for gamma_Cr_V")
    if (length(w) != dim(post$gamma_Cr)[3]) stop("crop weight length != Cr_n for gamma_Cr")

    cropV_draws <- post$gamma_Cr_V[draw_idx, , drop = FALSE]
    gamma_CrV_star <- as.numeric(cropV_draws %*% w)

    cropC_draws <- post$gamma_Cr[draw_idx, j_idx, , drop = FALSE]  # draws × 1 × Cr_n
    cropC_mat   <- matrix(cropC_draws, nrow = n_s, ncol = length(w))
    gamma_CrC_star <- as.numeric(cropC_mat %*% w)

    crop_label <- "avg_over_crops_weighted"
  } else {
    Cr_idx <- crop_lookup$Cr[crop_lookup$Crop == crop]
    if (length(Cr_idx) != 1L) stop("crop not found in crop_lookup (check exact spelling/levels)")
    gamma_CrV_star <- post$gamma_Cr_V[draw_idx, Cr_idx]
    gamma_CrC_star <- post$gamma_Cr[draw_idx, j_idx, Cr_idx]
    crop_label <- as.character(crop)
  }

  # ---------------- Residue handling ----------------
  if (!is.null(res_z) && (length(res_z) != 1L || !is.numeric(res_z))) {
    stop("res_z must be NULL or a single numeric value.")
  }

  if (is.null(res_z)) {
    res_prop <- RES_bar
    res_mode <- "avg_observed"
  } else {
    res_prop <- as.numeric(res_z)
    if (!is.finite(res_prop)) stop("res_z must be finite when not NULL.")
    if (identical(res_prop, 0)) res_prop <- 1e-4
    res_mode <- "fixed"
  }
  if (res_prop <= 0 || res_prop >= 1) stop("Residue proportion must be strictly between 0 and 1. Got: ", res_prop)

  # ---------------- Posterior pieces for this analyte ----------------
  alpha_j      <- post$alpha[draw_idx, j_idx]
  beta_stirj   <- post$beta_stir[draw_idx, j_idx]
  beta_cinj    <- post$beta_cin[draw_idx, j_idx]
  beta_dupj    <- post$beta_dup[draw_idx, j_idx]
  beta_resC_j  <- post$beta_res_C[draw_idx, j_idx]

  beta_vol     <- post$beta_vol[draw_idx]

  # v1p8: analyte-specific irrigation effect
  beta_irrj    <- post$beta_irr[draw_idx, j_idx]

  a_V          <- post$a_V[draw_idx]
  b_V          <- post$b_V[draw_idx]
  beta_res_V   <- post$beta_res_V[draw_idx]

  # sigmas
  sigma_V      <- post$sigma_V[draw_idx]
  sigma_VOLobs <- post$sigma_VOL_obs[draw_idx]
  sigma_Cproc  <- post$sigma_analyte[draw_idx]
  sigma_Cobs_j <- post$sigma_C_obs[draw_idx, j_idx]

  # GP on year
  if (is.null(year)) {
    F_sub  <- post$F_year[draw_idx, , j_idx, drop = FALSE]  # draws × Y × 1
    k_star <- apply(F_sub, 1, mean)
    year_label <- "avg_all_years"
  } else {
    Y_idx <- match(year, years_used)
    if (is.na(Y_idx)) stop("requested year not in years_used")
    k_star <- post$F_year[draw_idx, Y_idx, j_idx]
    year_label <- as.character(year)
  }

  # ---------------- Loop over STIR values ----------------
  n_x <- length(STIR_z_seq)

  # link-style (mean function) summaries
  load_mean <- numeric(n_x)
  load_low  <- numeric(n_x)
  load_high <- numeric(n_x)

  # posterior predictive summaries (latent + obs noise)
  load_pp_mean <- numeric(n_x)
  load_pp_low  <- numeric(n_x)
  load_pp_high <- numeric(n_x)

  for (k in seq_len(n_x)) {
    stz <- STIR_z_seq[k]
    RES_star <- rep(res_prop, n_s)

    # mean functions on z-scale
    muV_z <- a_V +
      b_V * stz +
      beta_res_V * RES_star +
      gamma_CrV_star

    muC_z <- alpha_j +
      beta_stirj * stz +
      beta_cinj  * cin_z +
      beta_vol   * muV_z +
      beta_irrj  * IRR_star_z +     # << v1p8 IRR_z scale
      beta_dupj  * DUP_star +
      beta_resC_j * RES_star +
      gamma_CrC_star +
      gamma_B_star +
      gamma_S_star +
      gamma_F_star +
      k_star

    # ---- (1) Link-style: mean relationship only ----
    # Keep the fitted latent model unchanged, but do not allow reported
    # concentration or volume predictions to fall below zero.
    C_raw_mu <- clamp_nonnegative(c_mean_j + c_sd_j * muC_z)
    V_raw_mu <- clamp_nonnegative(V_mean   + V_sd   * muV_z)
    L_link   <- C_raw_mu * V_raw_mu / 1000

    load_mean[k] <- mean(L_link)
    hpdi_k       <- HPDI(L_link, prob = ci_prob)
    load_low[k]  <- hpdi_k[1]
    load_high[k] <- hpdi_k[2]

    # ---- (2) Posterior predictive: latent + observation noise ----
    V_true_z <- rnorm(n_s, mean = muV_z, sd = sigma_V)
    V_obs_z  <- rnorm(n_s, mean = V_true_z, sd = sigma_VOLobs)

    C_true_z <- rnorm(n_s, mean = muC_z, sd = sigma_Cproc)
    C_obs_z  <- rnorm(n_s, mean = C_true_z, sd = sigma_Cobs_j)

    C_raw_pp <- clamp_nonnegative(c_mean_j + c_sd_j * C_obs_z)
    V_raw_pp <- clamp_nonnegative(V_mean   + V_sd   * V_obs_z)
    L_pp     <- C_raw_pp * V_raw_pp / 1000

    load_pp_mean[k] <- mean(L_pp)
    hpdi_pp         <- HPDI(L_pp, prob = ci_prob)
    load_pp_low[k]  <- hpdi_pp[1]
    load_pp_high[k] <- hpdi_pp[2]
  }

  data.frame(
    analyte            = analyte,
    STIR_raw           = STIR_raw_seq,

    # mean relationship (link-style)
    load_mean          = load_mean,
    load_low           = load_low,
    load_high          = load_high,

    # posterior predictive (sim-style)
    load_pp_mean       = load_pp_mean,
    load_pp_low        = load_pp_low,
    load_pp_high       = load_pp_high,

    block_level        = block_label,
    sampler_level      = sampler_label,
    flume_level        = flume_label,
    irrigation_level   = irrigation_label,
    duplicate_level    = duplicate_label,
    crop_level         = crop_label,
    residue_mode       = res_mode,
    res_z              = if (is.null(res_z)) NA_real_ else as.numeric(res_z),
    res_fixed_prop     = res_prop,
    cin_z              = cin_z,
    year_level         = year_label,
    ci_prob            = ci_prob
  )
}

simulate_load_curves <- function(
  post,
  analytes     = NULL,
  STIR_raw_seq = seq(0, 300, length.out = 60),
  block        = NULL,
  sampler      = NULL,
  flume        = NULL,
  irrigation   = NULL,
  duplicate    = NULL,
  crop         = NULL,
  res_z        = NULL,
  cin_z        = 0,
  year         = NULL,
  n_draws      = 2000,
  ci_prob      = 0.95
) {
  if (is.null(analytes)) analytes <- as.character(analyte_lookup$analyte_abbr)

  res_list <- lapply(analytes, function(an) {
    simulate_load_curve_single(
      post         = post,
      analyte      = an,
      STIR_raw_seq = STIR_raw_seq,
      block        = block,
      sampler      = sampler,
      flume        = flume,
      irrigation   = irrigation,
      duplicate    = duplicate,
      crop         = crop,
      res_z        = res_z,
      cin_z        = cin_z,
      year         = year,
      n_draws      = n_draws,
      ci_prob      = ci_prob
    )
  })

  dplyr::bind_rows(res_list)
}

compute_observed_load_points <- function(
  d_mod,
  analytes,
  block = NULL,
  sampler = NULL,
  flume = NULL,
  irrigation = NULL,
  duplicate = NULL,
  crop = NULL,
  year = NULL
) {

  df <- d_mod

  # Ensure analyte label exists
  if (!("analyte_abbr" %in% names(df))) {
    if ("A" %in% names(df)) {
      df$analyte_abbr <- analyte_lookup$analyte_abbr[match(df$A, analyte_lookup$A)]
    } else {
      stop("compute_observed_load_points(): need d_mod$analyte_abbr or d_mod$A.")
    }
  }

  # Filter analytes
  df <- df[df$analyte_abbr %in% analytes, , drop = FALSE]

  # Optional filters (apply only if columns exist)
  if (!is.null(block)      && "Rep" %in% names(df))           df <- df[df$Rep == block, , drop = FALSE]
  if (!is.null(sampler)    && "SampleMethod" %in% names(df))  df <- df[df$SampleMethod == sampler, , drop = FALSE]
  if (!is.null(flume)      && "FlumeMethod" %in% names(df))   df <- df[df$FlumeMethod == flume, , drop = FALSE]
  if (!is.null(irrigation) && "irr_num" %in% names(df))       df <- df[df$irr_num == irrigation, , drop = FALSE]
  if (!is.null(duplicate)  && "dup" %in% names(df))           df <- df[df$dup == as.numeric(duplicate), , drop = FALSE]

  if (!is.null(crop)) {
    if ("Crop" %in% names(df)) df <- df[df$Crop == crop, , drop = FALSE]
    if ("Cr" %in% names(df)) {
      Cr_idx <- crop_lookup$Cr[crop_lookup$Crop == crop][1]
      df <- df[df$Cr == Cr_idx, , drop = FALSE]
    }
  }

  if (!is.null(year)) {
    if ("year" %in% names(df)) df <- df[df$year == year, , drop = FALSE]
    if ("Year" %in% names(df)) df <- df[df$Year == year, , drop = FALSE]
  }

  req <- c("Result_mg_L", "Volume", "Season_STIR_toDate", "analyte_abbr")
  miss <- setdiff(req, names(df))
  if (length(miss) > 0) stop("compute_observed_load_points(): missing: ", paste(miss, collapse = ", "))

  df <- df[is.finite(df$Result_mg_L) & is.finite(df$Volume) & is.finite(df$Season_STIR_toDate), , drop = FALSE]

  df$STIR_raw <- df$Season_STIR_toDate
  df$load_obs <- df$Result_mg_L * df$Volume / 1000  # mg -> g

  # CRITICAL: provide the faceting variable used by predictions
  df$analyte <- as.character(df$analyte_abbr)

  df
}

simulate_load_curves_with_obs <- function(
  post,
  d_mod,
  analytes     = NULL,
  STIR_raw_seq = seq(0, 300, length.out = 60),
  block        = NULL,
  sampler      = NULL,
  flume        = NULL,
  irrigation   = NULL,
  duplicate    = NULL,
  crop         = NULL,
  res_z        = NULL,
  cin_z        = 0,
  year         = NULL,
  n_draws      = 2000,
  ci_prob      = 0.95
) {

  pred <- simulate_load_curves(
    post         = post,
    analytes     = analytes,
    STIR_raw_seq = STIR_raw_seq,
    block        = block,
    sampler      = sampler,
    flume        = flume,
    irrigation   = irrigation,
    duplicate    = duplicate,
    crop         = crop,
    res_z        = res_z,
    cin_z        = cin_z,
    year         = year,
    n_draws      = n_draws,
    ci_prob      = ci_prob
  )

  obs <- compute_observed_load_points(
    d_mod      = d_mod,
    analytes   = unique(pred$analyte),
    block      = block,
    sampler    = sampler,
    flume      = flume,
    irrigation = irrigation,
    duplicate  = duplicate,
    crop       = crop,
    year       = year
  )

  list(pred = pred, obs = obs)
}
```


## Scenario-based STIR → Load Simulation Functions

### Parameter definitions and behavior

The simulation functions allow you to generate posterior predicted analyte loads across a range of STIR values under different user-defined “scenarios.” Each scenario variable may be explicitly specified or left blank. When a parameter is left blank (NULL), the model automatically averages over that variable by using the posterior mean of its random effect or the empirical mean of the covariate.

**analyte**

* Character analyte abbreviation (for example, `"TP"`, `"NO3"`, `"TDS"`).
* Default behavior: uses *all analytes*.
* Valid levels come from `analyte_lookup$analyte_abbr`:
  NH4, ICP, NO3, NOx, NO2, NPOC, OP, Se, TDS, TKN, TN, TP, TSP, TSS.

**block**

* Field block or replication factor, matching `block_lookup$Rep` (typically `"1"` or `"2"`).
* If left NULL: averaged over blocks via the posterior mean of the random effect `gamma_B`.

**sampler**

* Sampler method type, matching `sampler_lookup$SampleMethod` (GB, GB3, GBH, GBW, ISC, LC).
* If left NULL: averaged over sampler methods via the posterior mean of `gamma_S`.

**flume**

* Flume hardware type, matching `flume_lookup$FlumeMethod`:
  “10 V”, “60 V Trap”, “7 V”, “8 V”, “Weir”.
* If left NULL: averaged using the posterior mean of `gamma_F`.

**irrigation**

* Irrigation-event identity, matching `irrigation_lookup$Irrigation`:
  1, 10, 2, 3, 4, 5, 6, 7, 8, 9, S1, S2.
* IRR is a fixed-effect covariate, not a random effect.
* If left NULL: the function uses the empirical mean of `irr_num` in the dataset.
* v1p8: internally converted to standardized IRR_z to match the Stan model.

**duplicate**

* Laboratory duplicate indicator (TRUE or FALSE).
* If NULL: uses the empirical mean duplicate rate across the dataset.

**crop**

* Character crop name matching `crop_lookup$Crop`: `"barley"`, `"dry beans"`, `"grain corn"`, `"silage corn"`, `"winter wheat"`.
* Affects both volume (`gamma_Cr_V`) and concentration (`gamma_Cr`) random effects.
* If NULL: averaged over crops using empirical frequency weights from `d_mod`.

**res_z**

* Residue proportion used as an input covariate in both volume and concentration models.
* Must be strictly between 0 and 1 (interpreted as proportion, not percent).
* If NULL: uses the empirical mean residue proportion from `d_mod`.
* The residue submodel is not simulated here, residue is treated as a fixed input scenario.

**cin_z**

* Standardized inflow concentration for the selected analyte.
* Defaults to 0 (“average inflow concentration”).

**STIR_raw_seq**

* The raw STIR values over which predictions are generated (for example, 0–300).

**n_draws**

* Number of posterior samples used for simulation, for computational efficiency.

**ci_prob**

* Credible interval probability for load estimates (default 0.95).

---


### Output

The functions return a tidy data frame containing:

* analyte
* STIR_raw
* load_mean (posterior mean load in mg)
* load_low (lower HPDI)
* load_high (upper HPDI)
* load_pp_mean (posterior predictive mean load)
* load_pp_low
* load_pp_high

This output is designed for straightforward faceted visualization in `ggplot2`.

```{r load-generative-curves-user-guide}
###############################################################
## USER INPUT GUIDE FOR `simulate_load_curves_with_obs()` (v1p8)
###############################################################

# ======================================================================
# Example scenario run (v1p8 dual ribbons)
# ======================================================================

res_scenario <- simulate_load_curves_with_obs(
  post         = post,
  d_mod        = d_mod,
  analytes     = NULL,
  STIR_raw_seq = seq(0, 250, length.out = 60),
  block        = NULL,
  sampler      = NULL,
  flume        = NULL,
  irrigation   = NULL,
  duplicate    = FALSE,
  crop         = NULL,
  res_z        = 0.10,   # FIXED residue proportion (0.10 = 10% cover). Use NULL to average observed residue.
  cin_z        = 0,
  year         = NULL,
  n_draws      = 2000,
  ci_prob      = 0.95
)

# res_scenario <- simulate_load_curves_with_obs(
#   post         = post,
#   d_mod        = d_mod,
#   analytes     = NULL,
#   STIR_raw_seq = seq(0, 300, length.out = 60),
#   block        = 1,
#   sampler      = NULL,
#   flume        = "60 V Trap",
#   irrigation   = 1,
#   duplicate    = FALSE,
#   crop         = "silage corn",
#   res_z        = .15,   # FIXED residue proportion (0.10 = 10% cover). Use NULL to average observed residue.
#   cin_z        = 0,
#   year         = 2023,
#   n_draws      = 2000,
#   ci_prob      = 0.95
# )

# res_scenario <- simulate_load_curves_with_obs(
#   post         = post,
#   d_mod        = d_mod,
#   analytes     = NULL,
#   STIR_raw_seq = seq(0, 300, length.out = 60),
#   block        = 1,
#   sampler      = NULL,
#   flume        = "7 V",
#   irrigation   = NULL,
#   duplicate    = FALSE,
#   crop         = "barley",
#   res_z        = .15,   # FIXED residue proportion (0.10 = 10% cover). Use NULL to average observed residue.
#   cin_z        = 0,
#   year         = 2016,
#   n_draws      = 2000,
#   ci_prob      = 0.95
# )

load_pred_scenario <- res_scenario$pred
load_obs_scenario  <- res_scenario$obs

# Build a dynamic subtitle from scenario labels
scenario_info <- load_pred_scenario %>%
  dplyr::distinct(
    year_level,
    block_level, sampler_level, flume_level,
    irrigation_level, duplicate_level,
    crop_level, residue_mode, res_fixed_prop, cin_z
  )

residue_text <- if (scenario_info$residue_mode == "avg_observed") {
  sprintf("avg observed %.1f%%", 100 * scenario_info$res_fixed_prop)
} else {
  sprintf("fixed %.1f%%", 100 * scenario_info$res_fixed_prop)
}

subtitle_text <- sprintf(
  "Year: %s; Block: %s; Sampler: %s;\nFlume: %s; Irrigation: %s;\nDuplicate: %s; Crop: %s;\nResidue: %s; Inflow(z)=%.2f",
  scenario_info$year_level,
  scenario_info$block_level,
  scenario_info$sampler_level,
  scenario_info$flume_level,
  scenario_info$irrigation_level,
  scenario_info$duplicate_level,
  scenario_info$crop_level,
  residue_text,
  scenario_info$cin_z
)

ci_pct <- round(100 * unique(load_pred_scenario$ci_prob)[1])

# Dual-ribbon plot:
#   - outer ribbon (pp) = latent process + observation noise
#   - inner ribbon (mean) = uncertainty in expected relationship
p_load <- ggplot(load_pred_scenario, aes(x = STIR_raw)) +
  geom_ribbon(
    aes(ymin = load_pp_low, ymax = load_pp_high),
    fill = "gray75", alpha = 0.35, linewidth = 0
  ) +
  geom_ribbon(
    aes(ymin = load_low, ymax = load_high),
    fill = "gray45", alpha = 0.35, linewidth = 0
  ) +
  geom_line(aes(y = load_mean), linewidth = 1.1) +
  geom_point(
    data = load_obs_scenario,
    aes(x = STIR_raw, y = load_obs, group = analyte),
    inherit.aes = FALSE,
    size = 2.0, alpha = 0.9
  ) +
  facet_wrap(~ analyte, scales = "free_y", ncol = 4) +
  labs(
    x = "Seasonal STIR",
    y = sprintf("Load (g per event), ribbons show %d%% intervals", ci_pct),
    title = "STIR → Load curves by analyte",
    subtitle = paste0(
      subtitle_text,
      "\nDark ribbon: mean relationship uncertainty. Light ribbon: posterior predictive (process + observation)."
    )
  ) +
  theme_classic(base_size = 18) +
  theme(
    strip.text    = element_text(size = 16, face = "bold"),
    axis.text     = element_text(size = 14),
    axis.title    = element_text(size = 18),
    plot.title    = element_text(size = 22, face = "bold"),
    plot.subtitle = element_text(size = 14),
    panel.spacing = unit(1.2, "lines"),
    plot.margin   = margin(15, 20, 15, 20)
  )

p_load
```


```{r save load fig}
# Save to figs folder (v1p8)
ggsave(
  filename = "../figs/load1p8_STIR_load_curves.jpeg",
  plot = p_load,
  width = 12, height = 10, dpi = 300
)
```

Plot effects of time (i.e., Year)
```{r year covar plot}
# ============================================================
# Year-distance covariance functions (prior vs posterior)
# Minimal-change version for v1p8
# Preserves original plotting behavior
# ============================================================

# ---- Safety checks ----
stopifnot(exists("post"))

# cov_fun helper, if not already defined
if (!exists("cov_fun")) {
  cov_fun <- function(etasq, rhosq, d) {
    etasq <- as.numeric(etasq)
    rhosq <- as.numeric(rhosq)
    d     <- as.numeric(d)
    etasq * exp(-rhosq * d^2)
  }
}

# transparent color helper, if not already defined
if (!exists("col_alpha")) {
  col_alpha <- function(col, alpha = 1) {
    grDevices::adjustcolor(col, alpha.f = alpha)
  }
}

# ---- Build year support from available data/post object ----
# Prefer explicit years_used if you already created it upstream
if (!exists("years_used")) {
  if (exists("dat_used") && !is.null(dat_used$year)) {
    years_used <- sort(unique(as.numeric(dat_used$year)))
  } else if (exists("year_id") && exists("year_lookup")) {
    years_used <- as.numeric(year_lookup[sort(unique(year_id))])
    years_used <- sort(unique(years_used))
  } else {
    # last-resort fallback: use sequential spacing across Y_n
    Y_n_local <- dim(post$F_year)[2]
    years_used <- seq_len(Y_n_local)
    warning("years_used was not found, using sequential indices 1:Y_n as fallback.")
  }
}

years_used <- sort(unique(as.numeric(years_used)))
years_used <- years_used[is.finite(years_used)]

if (length(years_used) < 2) {
  Y_n_local <- dim(post$F_year)[2]
  years_used <- seq_len(Y_n_local)
  warning("years_used had fewer than 2 finite values, using sequential indices 1:Y_n.")
}

# Distance support for x-axis
d_max <- max(years_used) - min(years_used)
if (!is.finite(d_max) || d_max <= 0) {
  d_max <- max(1, length(years_used) - 1)
  warning("Computed d_max was non-finite or <= 0, using index-based fallback.")
}

x_seq <- seq(0, d_max, length.out = 200)

# ---- Prior draws ----
# Keep the original prior simulation exactly as before
n_prior <- 50
set.seed(1)
etasq_prior <- rexp(n_prior, rate = 2)
rhosq_prior <- rexp(n_prior, rate = 0.5)

# ---- Posterior draws ----
# Pull directly from post
etasq_all <- as.numeric(post$etasq_year)
rhosq_all <- as.numeric(post$rhosq_year)

# Drop non-finite values, just in case
ok_post <- is.finite(etasq_all) & is.finite(rhosq_all)
etasq_all <- etasq_all[ok_post]
rhosq_all <- rhosq_all[ok_post]

if (length(etasq_all) == 0 || length(rhosq_all) == 0) {
  stop("No valid posterior GP hyperparameter draws found in post$etasq_year / post$rhosq_year.")
}

# Subsample posterior curves for the cloud
# Keep the original logic exactly
n_post <- min(50, length(etasq_all))
set.seed(2)
idx_post <- sample(seq_along(etasq_all), size = n_post, replace = FALSE)
etasq_post <- etasq_all[idx_post]
rhosq_post <- rhosq_all[idx_post]

# ---- Mean curves: prior and posterior ----
# Keep original behavior:
# prior mean from all prior draws
# posterior mean from all posterior draws
prior_cov <- vapply(
  x_seq,
  FUN = function(x) mean(cov_fun(etasq_prior, rhosq_prior, x), na.rm = TRUE),
  FUN.VALUE = numeric(1)
)
prior_mean <- prior_cov

post_cov <- vapply(
  x_seq,
  FUN = function(x) mean(cov_fun(etasq_all, rhosq_all, x), na.rm = TRUE),
  FUN.VALUE = numeric(1)
)
post_mean <- post_cov

# y-axis upper bound
y_max <- max(
  cov_fun(etasq_prior, rhosq_prior, 0),
  cov_fun(etasq_all,  rhosq_all,  0),
  prior_mean,
  post_mean,
  na.rm = TRUE
)

# ---------------------------------
# Styling choices
# ---------------------------------
col_prior_mean <- "black"
col_post_mean  <- "red"
lty_prior_mean <- 3
lty_post_mean  <- 2

col_prior_draw <- col_alpha("grey10", 0.18)
col_post_draw  <- col_alpha("firebrick", 0.25)

# ---------------------------------
# Reusable plotting function
# ---------------------------------
draw_year_cov_plot <- function() {
  plot(
    NULL,
    xlab = "distance (years)",
    ylab = "covariance",
    xlim = c(0, d_max-8),
    ylim = c(0, y_max*.9),
    main = "Year GP covariance: prior vs posterior"
  )

  # PRIOR curves
  for (i in seq_len(n_prior)) {
    curve(
      cov_fun(etasq_prior[i], rhosq_prior[i], x),
      from = 0, to = d_max,
      add = TRUE,
      col = col_prior_draw,
      lwd = 2
    )
  }

  # POSTERIOR curves
  for (i in seq_len(n_post)) {
    curve(
      cov_fun(etasq_post[i], rhosq_post[i], x),
      from = 0, to = d_max,
      add = TRUE,
      col = col_post_draw,
      lwd = 2
    )
  }

  # Mean lines
  lines(x_seq, prior_mean, col = col_prior_mean, lwd = 4, lty = lty_prior_mean)
  lines(x_seq, post_mean,  col = col_post_mean,  lwd = 4, lty = lty_post_mean)

  legend(
    "topright",
    legend = c("Prior draws", "Posterior draws", "Prior mean", "Posterior mean"),
    col    = c(col_prior_draw, col_post_draw, col_prior_mean, col_post_mean),
    lwd    = c(2, 2, 4, 4),
    lty    = c(1, 1, lty_prior_mean, lty_post_mean),
    bty    = "n"
  )
}

# ---------------------------------
# Save figure
# ---------------------------------
jpeg("../figs/year_covariance_gp_v1p8.jpeg", width = 2000, height = 1500, res = 220)
draw_year_cov_plot()
dev.off()

# ---------------------------------
# Also display inline
# ---------------------------------
draw_year_cov_plot()

```


```{r analyte persistence plot}
## ============================================================
## Analyte persistence: lag-1 correlation of latent year effect
## v1p8 (uses post$F_year: draws × Y × A)
## Ridgeline replacement (overwrites same filename pattern)
## ============================================================

F_arr <- post$F_year
stopifnot(length(dim(F_arr)) == 3)

n_draws <- dim(F_arr)[1]
Y_n     <- dim(F_arr)[2]
A_n     <- dim(F_arr)[3]

# compute lag-1 correlations for each analyte and draw
lag1_mat <- matrix(NA_real_, nrow = n_draws, ncol = A_n)

for (a in seq_len(A_n)) {
  Fa <- F_arr[, , a]  # draws × Y
  for (s in seq_len(n_draws)) {
    z <- Fa[s, ]
    if (is.finite(sd(z)) && sd(z) > 0) {
      lag1_mat[s, a] <- stats::cor(z[-Y_n], z[-1])
    }
  }
}

# ---- Build long posterior draw table (no forcats) ----
lag1_df <- as.data.frame(lag1_mat)
colnames(lag1_df) <- as.character(analyte_lookup$analyte_abbr)

lag1_long <- lag1_df %>%
  pivot_longer(
    cols = everything(),
    names_to = "analyte_abbr",
    values_to = "lag1_cor"
  ) %>%
  filter(is.finite(lag1_cor))

# order analytes by posterior mean lag-1 correlation
analyte_order <- lag1_long %>%
  group_by(analyte_abbr) %>%
  summarise(mu = mean(lag1_cor), .groups = "drop") %>%
  arrange(mu) %>%
  pull(analyte_abbr)

lag1_long$analyte_abbr <- factor(lag1_long$analyte_abbr, levels = analyte_order)

# ---- Ridgeline plot ----
p_persist_ridge <- ggplot(lag1_long, aes(x = lag1_cor, y = analyte_abbr)) +
  geom_vline(xintercept = 0, linetype = 2, linewidth = 1) +
  stat_density_ridges(
    fill = "darkred",
    alpha = 0.55,
    scale = 1.1,
    rel_min_height = 0.01,
    quantile_lines = TRUE,
    quantiles = 2,
    quantile_color = "black",
    quantile_linetype = 2,
    quantile_linewidth = 1,
    bandwidth = 0.02
  ) +
  labs(
    x = "Lag-1 correlation (F_year)",
    y = "Analyte (ordered by mean persistence)",
    title = "Analyte-specific persistence of temporal deviations"
  ) +
  theme_ridges(grid = TRUE, center_axis_labels = TRUE) +
  theme(
    legend.position = "none",
    text = element_text(size = 14),
    axis.text.y = element_text(size = 10)
  )

p_persist_ridge

# Save to figs folder
ggsave(
  filename = "../figs/analyte_persistence_precis_v1p8.jpeg",
  plot = p_persist_ridge,
  width = 12,
  height = 10,
  dpi = 300
)

```


```{r year effects plot}
## ============================================================
## Year effects plot: F_year by analyte (v1p8)
## post$F_year is draws × Y × A
## ============================================================

F_arr <- post$F_year
stopifnot(length(dim(F_arr)) == 3)

Y_n <- dim(F_arr)[2]
stopifnot(length(years_used) == Y_n)

A_n <- dim(F_arr)[3]

# posterior means and intervals for F_year
F_mean <- apply(F_arr, c(2, 3), mean, na.rm = TRUE)             # Y × A
F_hpdi <- apply(F_arr, c(2, 3), HPDI, prob = 0.89)              # 2 × Y × A

# Build long data frame
F_df <- expand.grid(
  Y = seq_len(Y_n),
  A = seq_len(A_n)
) %>%
  dplyr::mutate(
    effect_mean = as.vector(F_mean),          # length Y*A
    hpdi_low    = as.vector(F_hpdi[1, , ]),   # length Y*A
    hpdi_high   = as.vector(F_hpdi[2, , ])    # length Y*A
  ) %>%
  dplyr::left_join(analyte_lookup, by = "A") %>%
  dplyr::mutate(Year = years_used[Y])

gg_F <- ggplot(
  F_df,
  aes(x = Year, y = effect_mean, ymin = hpdi_low, ymax = hpdi_high)
) +
  geom_hline(yintercept = 0, lty = 2) +
  geom_ribbon(alpha = 0.25) +
  geom_line(linewidth = 0.8) +
  facet_wrap(~ analyte_abbr, scales = "free_y", ncol = 4) +
  labs(
    x = "Year",
    y = "Latent temporal deviation (z concentration scale)",
    title = "Year-to-year latent deviations by analyte"
  ) +
  theme_classic(base_size = 14)

gg_F

# Save to figs folder
ggsave(
  filename = "../figs/yearly_latent_deviations_v1p8.jpeg",
  plot = gg_F,
  width = 12, height = 10, dpi = 300
)

```


## Annual Load Table and Plot

```{r pred data generation}
# ============================================================
# v1p8: Build a prediction frame that includes "missing chemistry years"
# Drop-in replacement with:
#   1) robust irrigation mapping (keeps S1/S2, avoids coercion warnings)
#   2) diagnostics explaining why years/rows get dropped
#   3) optional cin_z fill (default: NA cin_z -> 0)
#   4) fields: Crop + residue placeholder (RES_z)
#      - Crop is required for Stan (Cr index)
#      - Residue is required for Stan (RES predictor)
#        For now we set RES_z = 0 (placeholder; no impact)
#
# IMPORTANT v1p8 note:
#   - Irrigation is standardized later as irr_z to match the Stan data scale.
# ============================================================
build_d_pred <- function(
  wq_clean,
  event_id_col = "orig_row",
  keep_outflow_only = TRUE,
  drop_no_runoff = TRUE,
  cin_na_to_zero = TRUE,
  diagnostics = TRUE,
  res_na_to_zero = TRUE  # if RES_z ever exists, fill NA -> 0
) {

  # required columns
  needed <- c(
    "Year","Treatment","analyte_abbr","Season_STIR_toDate",
    "Rep","SampleMethod","FlumeMethod","Irrigation","Duplicate","cin_z",
    "Crop",
    event_id_col
  )
  miss <- setdiff(needed, names(wq_clean))
  if (length(miss) > 0) stop("wq_clean missing columns: ", paste(miss, collapse = ", "))

  d0 <- wq_clean

  if (diagnostics) {
    message("\n---- Year coverage in raw wq_clean (before any filters) ----")
    print(wq_clean %>% dplyr::count(Year) %>% dplyr::arrange(Year))
  }

  # ------------------------------------------------------------
  # Filters (apply BEFORE transmute)
  # ------------------------------------------------------------
  if (keep_outflow_only && "InflowOutflow" %in% names(d0)) {
    d0 <- d0 %>% dplyr::filter(.data$InflowOutflow == "OUT")
  }

  if (drop_no_runoff && "NoRunoff" %in% names(d0)) {
    d0 <- d0 %>% dplyr::filter(!isTRUE(.data$NoRunoff))
  }

  if (diagnostics) {
    message("\n---- Year coverage after keep_outflow_only / drop_no_runoff filters ----")
    print(d0 %>% dplyr::count(Year) %>% dplyr::arrange(Year))

    message("\n---- Years removed by filters (raw vs filtered) ----")
    raw_years  <- wq_clean %>% dplyr::distinct(Year)
    kept_years <- d0       %>% dplyr::distinct(Year)
    removed <- dplyr::anti_join(raw_years, kept_years, by = "Year") %>% dplyr::arrange(Year)
    print(removed)
  }

  # ------------------------------------------------------------
  # Robust irrigation mapping (no "NAs introduced by coercion")
  # ------------------------------------------------------------
  irr_levels <- sort(unique(as.character(d0$Irrigation)))

  irr_map <- tibble::tibble(Irrigation_chr = irr_levels) %>%
    dplyr::mutate(
      irr_num = dplyr::case_when(
        Irrigation_chr == "S1" ~ 11,
        Irrigation_chr == "S2" ~ 12,
        TRUE ~ readr::parse_number(Irrigation_chr)
      )
    )

  if (diagnostics) {
    bad_irr <- irr_map %>% dplyr::filter(is.na(irr_num))
    if (nrow(bad_irr) > 0) {
      message("Unmapped Irrigation levels (these rows will be dropped): ",
              paste(bad_irr$Irrigation_chr, collapse = ", "))
    } else {
      message("All Irrigation levels mapped successfully (including S1/S2 if present).")
    }
  }

  d1 <- d0 %>%
    dplyr::mutate(Irrigation_chr = as.character(Irrigation)) %>%
    dplyr::left_join(irr_map, by = "Irrigation_chr")

  # ------------------------------------------------------------
  # Residue placeholder (Stan requires RES predictor)
  # Use an epsilon inside (0,1) as the boundary holder, never 0.
  # If a residue column exists, clamp it into (eps, 1-eps).
  # If missing entirely, create eps as a placeholder.
  # ------------------------------------------------------------
  eps_res <- 1e-4

  clamp01 <- function(x, eps = 1e-4) {
    x <- as.numeric(x)
    x[!is.finite(x)] <- NA_real_
    pmin(pmax(x, eps), 1 - eps)
  }

  if ("RES_z" %in% names(d1)) {
    d1 <- d1 %>%
      dplyr::mutate(RES_z = dplyr::if_else(is.na(RES_z), eps_res, clamp01(RES_z, eps_res)))
  } else if ("residue_z" %in% names(d1)) {
    d1 <- d1 %>%
      dplyr::mutate(RES_z = dplyr::if_else(is.na(.data$residue_z), eps_res, clamp01(.data$residue_z, eps_res)))
  } else if ("RES_input" %in% names(d1)) {
    d1 <- d1 %>%
      dplyr::mutate(RES_z = dplyr::if_else(is.na(.data$RES_input), eps_res, clamp01(.data$RES_input, eps_res)))
  } else if ("RES" %in% names(d1)) {
    d1 <- d1 %>%
      dplyr::mutate(RES_z = dplyr::if_else(is.na(.data$RES), eps_res, clamp01(.data$RES, eps_res)))
  } else if ("residue_prop" %in% names(d1)) {
    d1 <- d1 %>%
      dplyr::mutate(RES_z = dplyr::if_else(is.na(.data$residue_prop), eps_res, clamp01(.data$residue_prop, eps_res)))
  } else {
    d1 <- d1 %>% dplyr::mutate(RES_z = eps_res)
  }

  # ------------------------------------------------------------
  # Diagnostics BEFORE final dropping
  # ------------------------------------------------------------
  if (diagnostics) {

    na_irr_by_year <- d1 %>%
      dplyr::summarise(
        .by = Year,
        n = dplyr::n(),
        n_irr_na = sum(is.na(irr_num)),
        frac_irr_na = n_irr_na / n
      ) %>%
      dplyr::arrange(dplyr::desc(frac_irr_na))

    message("\n---- Diagnostic: NA irr_num by Year (top 10) ----")
    print(utils::head(na_irr_by_year, 10))

    diag_by_year <- d1 %>%
      dplyr::transmute(
        Year,
        na_event = is.na(.data[[event_id_col]]),
        na_stir  = is.na(Season_STIR_toDate),
        na_rep   = is.na(Rep),
        na_samp  = is.na(SampleMethod),
        na_flume = is.na(FlumeMethod),
        na_dup   = is.na(Duplicate),
        na_cin   = is.na(cin_z),
        na_crop  = is.na(Crop),
        na_res   = is.na(RES_z)
      ) %>%
      dplyr::summarise(
        .by = Year,
        n = dplyr::n(),
        frac_na_event = mean(na_event),
        frac_na_stir  = mean(na_stir),
        frac_na_rep   = mean(na_rep),
        frac_na_samp  = mean(na_samp),
        frac_na_flume = mean(na_flume),
        frac_na_dup   = mean(na_dup),
        frac_na_cin   = mean(na_cin),
        frac_na_crop  = mean(na_crop),
        frac_na_res   = mean(na_res)
      ) %>%
      dplyr::arrange(Year)

    message("\n---- Diagnostic: missing predictor fractions by Year ----")
    print(diag_by_year)
  }

  # ------------------------------------------------------------
  # Build final d_pred with model-ready columns
  # ------------------------------------------------------------
  d <- d1 %>%
    dplyr::transmute(
      event_id = .data[[event_id_col]],
      Year,
      Treatment,
      analyte_abbr,
      Crop,
      STIR_raw = Season_STIR_toDate,
      Rep,
      SampleMethod,
      FlumeMethod,
      irr_num,
      dup   = as.numeric(Duplicate),
      cin_z = if (cin_na_to_zero) dplyr::if_else(is.na(cin_z), 0, cin_z) else cin_z,
      RES_z = if (res_na_to_zero) dplyr::if_else(is.na(RES_z), 0, RES_z) else RES_z
    ) %>%
    dplyr::filter(
      !is.na(event_id),
      !is.na(Year),
      !is.na(Treatment),
      !is.na(analyte_abbr),
      !is.na(Crop),
      !is.na(STIR_raw),
      !is.na(Rep),
      !is.na(SampleMethod),
      !is.na(FlumeMethod),
      !is.na(irr_num),
      !is.na(dup),
      !is.na(cin_z),
      !is.na(RES_z)
    )

  if (diagnostics) {
    message("\n---- Final d_pred coverage by Year (after dropping incomplete rows) ----")
    print(d %>% dplyr::count(Year) %>% dplyr::arrange(Year))
  }

  d
}

# Run it
d_pred <- build_d_pred(wq_clean, event_id_col = "orig_row")

```


```{r real data generation}
make_annual_observed_table_all_treatments <- function(
  wq_clean,
  analyte_pick = NULL,
  keep_outflow_only = TRUE,
  drop_no_runoff = TRUE,
  B = 600,
  prob_ci = 0.95,
  seed = 1
) {
  needed <- c("Year","Treatment","analyte_abbr","Result_mg_L","Volume","orig_row")
  miss <- setdiff(needed, names(wq_clean))
  if (length(miss) > 0) stop("wq_clean missing columns: ", paste(miss, collapse = ", "))

  d0 <- wq_clean

  if (keep_outflow_only && "InflowOutflow" %in% names(d0)) {
    d0 <- d0 %>% dplyr::filter(.data$InflowOutflow == "OUT")
  }
  if (drop_no_runoff && "NoRunoff" %in% names(d0)) {
    d0 <- d0 %>% dplyr::filter(!isTRUE(.data$NoRunoff))
  }
  if (!is.null(analyte_pick)) {
    d0 <- d0 %>% dplyr::filter(as.character(.data$analyte_abbr) == analyte_pick)
  }

  # rows usable for observed load
  d_obs <- d0 %>%
    dplyr::filter(!is.na(.data$Result_mg_L), !is.na(.data$Volume)) %>%
    dplyr::mutate(
      Year = as.integer(.data$Year),
      analyte = as.character(.data$analyte_abbr),
      treatment = as.character(.data$Treatment),
      event_id = .data$orig_row,
      load_g = (.data$Result_mg_L * .data$Volume) / 1000,  # mg -> g
      vol_L  = .data$Volume
    ) %>%
    dplyr::select(Year, analyte, treatment, event_id, load_g, vol_L)

  q_lo <- (1 - prob_ci) / 2
  q_hi <- 1 - q_lo

  set.seed(seed)

  out <- d_obs %>%
    dplyr::group_by(.data$Year, .data$analyte, .data$treatment) %>%
    dplyr::group_modify(~{
      dat <- dplyr::distinct(.x, .data$event_id, .keep_all = TRUE)
      n_events <- nrow(dat)

      if (n_events == 0) {
        return(tibble::tibble(
          load_mean = NA_real_, load_low = NA_real_, load_high = NA_real_,
          volume_mean = NA_real_, volume_low = NA_real_, volume_high = NA_real_,
          conc_mean = NA_real_, conc_low = NA_real_, conc_high = NA_real_,
          n_events = 0L,
          prob_ci = prob_ci
        ))
      }

      # point estimates (annual totals)
      L_sum <- sum(dat$load_g, na.rm = TRUE)
      V_sum <- sum(dat$vol_L,  na.rm = TRUE)
      C_fw  <- if (is.finite(V_sum) && V_sum > 0) (L_sum * 1000 / V_sum) else NA_real_

      # If only one event, bootstrap is degenerate, CI == point estimate
      if (n_events == 1 || B <= 1) {
        return(tibble::tibble(
          load_mean = L_sum, load_low = L_sum, load_high = L_sum,
          volume_mean = V_sum, volume_low = V_sum, volume_high = V_sum,
          conc_mean = C_fw, conc_low = C_fw, conc_high = C_fw,
          n_events = as.integer(n_events),
          prob_ci = prob_ci
        ))
      }

      # bootstrap indices: n_events x B matrix
      idx_mat <- matrix(
        sample.int(n_events, size = n_events * B, replace = TRUE),
        nrow = n_events, ncol = B
      )

      load_mat <- matrix(dat$load_g[idx_mat], nrow = n_events, ncol = B)
      vol_mat  <- matrix(dat$vol_L[idx_mat],  nrow = n_events, ncol = B)

      Lb <- colSums(load_mat, na.rm = TRUE)
      Vb <- colSums(vol_mat,  na.rm = TRUE)
      Cb <- ifelse(Vb > 0, (Lb * 1000 / Vb), NA_real_)

      tibble::tibble(
        load_mean = L_sum,
        load_low  = stats::quantile(Lb, probs = q_lo, na.rm = TRUE, names = FALSE),
        load_high = stats::quantile(Lb, probs = q_hi, na.rm = TRUE, names = FALSE),

        volume_mean = V_sum,
        volume_low  = stats::quantile(Vb, probs = q_lo, na.rm = TRUE, names = FALSE),
        volume_high = stats::quantile(Vb, probs = q_hi, na.rm = TRUE, names = FALSE),

        conc_mean = C_fw,
        conc_low  = stats::quantile(Cb, probs = q_lo, na.rm = TRUE, names = FALSE),
        conc_high = stats::quantile(Cb, probs = q_hi, na.rm = TRUE, names = FALSE),

        n_events = as.integer(n_events),
        prob_ci = prob_ci
      )
    }) %>%
    dplyr::ungroup()

  out
}

# Example:
annual_tbl_obs <- make_annual_observed_table_all_treatments(
  wq_clean = wq_clean,
  B = 600,
  prob_ci = 0.95
)

```


```{r pred data table}
# ============================================================
# v1p8: Annual table from posterior predictions summed over d_pred
# IMPORTANT FIXES:
#   - De-duplicate to ONE row per event_id for each Year × Treatment × Analyte group before summing
#   - Irrigation is used on standardized scale irr_z (matches Stan data)
#   - beta_irr is analyte-specific in v1p8: post$beta_irr[draw, A]
# ============================================================
make_annual_load_table_all_treatments_pred <- function(
  post,
  d_pred,
  years_used,
  analyte_lookup,
  block_lookup,
  sampler_lookup,
  flume_lookup,
  crop_lookup,
  c_stats,
  V_mean, V_sd,
  stir_mean, stir_sd,
  irr_mean, irr_sd,             # NEW v1p8
  n_draws = 1500,
  prob_ci = 0.95,
  drop_unmapped_analytes = TRUE,
  drop_missing_c_stats = TRUE
) {
  needed_cols <- c(
    "event_id","Year","Treatment","analyte_abbr","STIR_raw",
    "Rep","SampleMethod","FlumeMethod","irr_num","dup","cin_z",
    "Crop","RES_z"
  )
  miss <- setdiff(needed_cols, names(d_pred))
  if (length(miss) > 0) stop("d_pred is missing columns: ", paste(miss, collapse = ", "))

  # draw subsample
  n_total <- nrow(post$alpha)
  draw_idx <- if (n_total > n_draws) sample.int(n_total, n_draws) else seq_len(n_total)
  D <- length(draw_idx)

  q_lo <- (1 - prob_ci) / 2
  q_hi <- 1 - q_lo
  ci <- function(x) stats::quantile(x, probs = c(q_lo, q_hi), na.rm = TRUE, names = FALSE)

  # map indices + stats
  dat <- d_pred %>%
    dplyr::left_join(analyte_lookup %>% dplyr::select(A, analyte_abbr), by = "analyte_abbr") %>%
    dplyr::left_join(block_lookup   %>% dplyr::select(B, Rep),          by = "Rep") %>%
    dplyr::left_join(sampler_lookup %>% dplyr::select(S, SampleMethod), by = "SampleMethod") %>%
    dplyr::left_join(flume_lookup   %>% dplyr::select(Fu, FlumeMethod), by = "FlumeMethod") %>%
    dplyr::left_join(crop_lookup    %>% dplyr::select(Cr, Crop),        by = "Crop") %>%
    dplyr::left_join(c_stats %>% dplyr::select(A, c_mean, c_sd) %>% dplyr::distinct(), by = "A") %>%
    dplyr::mutate(
      Y      = match(Year, years_used),
      STIR_z = (STIR_raw - stir_mean) / stir_sd,
      irr_z  = (irr_num  - irr_mean)  / irr_sd     # NEW v1p8
    )

  # ---- drop unmapped analytes instead of stopping ----
  if (anyNA(dat$A)) {
    bad_an <- sort(unique(dat$analyte_abbr[is.na(dat$A)]))
    msg <- paste0(
      "Some analyte_abbr not found in analyte_lookup: ",
      paste(bad_an, collapse = ", "),
      "."
    )
    if (!isTRUE(drop_unmapped_analytes)) {
      stop(msg)
    } else {
      warning(msg, " Dropping these rows from d_pred for annual table.")
      dat <- dat %>% dplyr::filter(!is.na(A))
    }
  }

  # ---- optionally drop rows missing c_stats ----
  if (anyNA(dat$c_mean) || anyNA(dat$c_sd)) {
    bad_an2 <- sort(unique(dat$analyte_abbr[is.na(dat$c_mean) | is.na(dat$c_sd)]))
    msg2 <- paste0(
      "Missing c_stats (c_mean/c_sd) for some analytes: ",
      paste(bad_an2, collapse = ", "),
      "."
    )
    if (!isTRUE(drop_missing_c_stats)) {
      stop(msg2)
    } else {
      warning(msg2, " Dropping these rows from d_pred for annual table.")
      dat <- dat %>% dplyr::filter(!is.na(c_mean), !is.na(c_sd))
    }
  }

  # Hard checks for structural lookups (these *should* be complete)
  if (anyNA(dat$B))  stop("Some Rep not found in block_lookup.")
  if (anyNA(dat$S))  stop("Some SampleMethod not found in sampler_lookup.")
  if (anyNA(dat$Fu)) stop("Some FlumeMethod not found in flume_lookup.")
  if (anyNA(dat$Cr)) stop("Some Crop not found in crop_lookup.")
  if (anyNA(dat$Y))  stop("Some Year not found in years_used.")
  if (anyNA(dat$irr_z)) stop("Some irr_z are NA. Check irr_mean/irr_sd and irr_num mapping.")

  # Integer indices for fast indexing
  dat <- dat %>%
    dplyr::mutate(
      A  = as.integer(A),
      B  = as.integer(B),
      S  = as.integer(S),
      Fu = as.integer(Fu),
      Y  = as.integer(Y),
      Cr = as.integer(Cr)
    )

  # Groups: all treatments automatically
  groups <- dat %>%
    dplyr::distinct(Year, analyte_abbr, Treatment) %>%
    dplyr::arrange(analyte_abbr, Year, Treatment)

  out <- vector("list", nrow(groups))

  for (g in seq_len(nrow(groups))) {

    yr <- groups$Year[g]
    an <- groups$analyte_abbr[g]
    tr <- groups$Treatment[g]

    dsub <- dat %>% dplyr::filter(Year == yr, analyte_abbr == an, Treatment == tr)
    if (nrow(dsub) == 0) next

    # de-duplicate analyte rows to one per event_id
    dsub1 <- dsub %>% dplyr::distinct(event_id, .keep_all = TRUE)

    # Row-level predictors (use dsub1)
    A  <- dsub1$A;  B <- dsub1$B;  S <- dsub1$S;  Fu <- dsub1$Fu;  Y <- dsub1$Y;  Cr <- dsub1$Cr
    stz <- dsub1$STIR_z
    cin <- dsub1$cin_z
    irz <- dsub1$irr_z              # NEW v1p8
    dup <- dsub1$dup
    res <- dsub1$RES_z
    c_mean <- dsub1$c_mean
    c_sd   <- dsub1$c_sd

    # Event-level ledger for volumes (also 1 row per event_id)
    ev <- dsub %>%
      dplyr::distinct(event_id, Year, Treatment, STIR_z, Crop, RES_z, Rep, SampleMethod, FlumeMethod, irr_num, irr_z, dup) %>%
      dplyr::left_join(crop_lookup    %>% dplyr::select(Cr, Crop),        by = "Crop") %>%
      dplyr::left_join(block_lookup   %>% dplyr::select(B, Rep),          by = "Rep") %>%
      dplyr::left_join(sampler_lookup %>% dplyr::select(S, SampleMethod), by = "SampleMethod") %>%
      dplyr::left_join(flume_lookup   %>% dplyr::select(Fu, FlumeMethod), by = "FlumeMethod") %>%
      dplyr::mutate(
        Cr = as.integer(Cr),
        B  = as.integer(B),
        S  = as.integer(S),
        Fu = as.integer(Fu)
      )

    stz_ev <- ev$STIR_z
    res_ev <- ev$RES_z
    Cr_ev  <- ev$Cr

    V_ann   <- numeric(D)
    L_ann_g <- numeric(D)
    C_fw    <- numeric(D)

    for (dd in seq_len(D)) {
      s <- draw_idx[dd]

      # ---- Predict volume per event (one per event_id) ----
      muV_z_ev <- post$a_V[s] +
        post$b_V[s]        * stz_ev +
        post$beta_res_V[s] * res_ev +
        post$gamma_Cr_V[s, Cr_ev]

      # Post-process volume predictions onto the physically admissible domain.
      V_ev_raw <- clamp_nonnegative(V_mean + V_sd * muV_z_ev)
      Vsum <- sum(V_ev_raw, na.rm = TRUE)

      # ---- Predict concentration per analyte event-row (one per event_id) ----
      muV_z_row <- post$a_V[s] +
        post$b_V[s]        * stz +
        post$beta_res_V[s] * res +
        post$gamma_Cr_V[s, Cr]

      gB <- post$gamma_B[s, , , drop = TRUE][cbind(A, B)]
      gS <- post$gamma_S[s, , , drop = TRUE][cbind(A, S)]
      gF <- post$gamma_F[s, , , drop = TRUE][cbind(A, Fu)]
      kY <- post$F_year[s, , , drop = TRUE][cbind(Y, A)]
      gCrC <- post$gamma_Cr[s, , , drop = TRUE][cbind(A, Cr)]

      muC_z <- post$alpha[s, A] +
        post$beta_stir[s, A]  * stz +
        post$beta_cin[s, A]   * cin +
        post$beta_vol[s]      * muV_z_row +
        post$beta_irr[s, A]   * irz +      # NEW v1p8: analyte-specific and standardized
        post$beta_dup[s, A]   * dup +
        post$beta_res_C[s, A] * res +
        gCrC +
        gB + gS + gF + kY

      # Post-process concentration predictions onto the physically admissible
      # domain before computing annual summaries and loads.
      C_raw <- clamp_nonnegative(c_mean + c_sd * muC_z)

      # ---- Annual load for this analyte: sum across events ONCE ----
      V_row_raw <- clamp_nonnegative(V_mean + V_sd * muV_z_row)
      Lmg <- sum(C_raw * V_row_raw, na.rm = TRUE)
      Lg  <- Lmg / 1000

      V_ann[dd]   <- Vsum
      L_ann_g[dd] <- Lg
      C_fw[dd]    <- if (is.finite(Vsum) && Vsum > 0) ((Lg * 1000) / Vsum) else NA_real_
    }

    out[[g]] <- tibble::tibble(
      Year = yr,
      analyte = an,
      treatment = tr,
      volume_mean   = clamp_nonnegative(mean(V_ann,   na.rm = TRUE)),
      volume_low    = clamp_nonnegative(ci(V_ann)[1]),
      volume_high   = clamp_nonnegative(ci(V_ann)[2]),
      conc_mean     = clamp_nonnegative(mean(C_fw,    na.rm = TRUE)),
      conc_low      = clamp_nonnegative(ci(C_fw)[1]),
      conc_high     = clamp_nonnegative(ci(C_fw)[2]),
      load_mean     = clamp_nonnegative(mean(L_ann_g, na.rm = TRUE)),
      load_low      = clamp_nonnegative(ci(L_ann_g)[1]),
      load_high     = clamp_nonnegative(ci(L_ann_g)[2]),
      n_events      = nrow(ev),
      n_rows        = nrow(dsub1),
      prob_ci       = prob_ci
    )
  }

  dplyr::bind_rows(out) %>%
    dplyr::arrange(analyte, Year, treatment)
}

# Example usage:
annual_tbl_pred <- make_annual_load_table_all_treatments_pred(
  post = post,
  d_pred = d_pred,
  years_used = years_used,
  analyte_lookup = analyte_lookup,
  block_lookup = block_lookup,
  sampler_lookup = sampler_lookup,
  flume_lookup = flume_lookup,
  crop_lookup = crop_lookup,
  c_stats = c_stats,
  V_mean = V_mean, V_sd = V_sd,
  stir_mean = stir_mean, stir_sd = stir_sd,
  irr_mean = irr_mean, irr_sd = irr_sd
)

```


```{r bind obs and modeled}
bind_observed_modeled <- function(annual_tbl_pred, annual_tbl_obs) {

  # ----------------------------
  # Helper: normalize CI columns to *_low95/*_high95
  # ----------------------------
  normalize_ci <- function(df, prefix, source_label) {

    mean_nm <- paste0(prefix, "_mean")
    low95_nm  <- paste0(prefix, "_low95")
    high95_nm <- paste0(prefix, "_high95")
    low_nm    <- paste0(prefix, "_low")
    high_nm   <- paste0(prefix, "_high")

    if (all(c(mean_nm, low95_nm, high95_nm) %in% names(df))) return(df)

    if (all(c(mean_nm, low_nm, high_nm) %in% names(df))) {
      df[[low95_nm]]  <- df[[low_nm]]
      df[[high95_nm]] <- df[[high_nm]]
      return(df)
    }

    stop(
      sprintf(
        "%s table is missing CI columns for '%s'. Expected either (%s,%s,%s) or (%s,%s,%s). Present: %s",
        source_label, prefix,
        mean_nm, low95_nm, high95_nm,
        mean_nm, low_nm, high_nm,
        paste(names(df), collapse = ", ")
      )
    )
  }

  # ---- normalize both inputs ----
  annual_tbl_pred <- normalize_ci(annual_tbl_pred, "load",   "Modeled")
  annual_tbl_pred <- normalize_ci(annual_tbl_pred, "conc",   "Modeled")
  annual_tbl_pred <- normalize_ci(annual_tbl_pred, "volume", "Modeled")

  annual_tbl_obs  <- normalize_ci(annual_tbl_obs,  "load",   "Observed")
  annual_tbl_obs  <- normalize_ci(annual_tbl_obs,  "conc",   "Observed")
  annual_tbl_obs  <- normalize_ci(annual_tbl_obs,  "volume", "Observed")

  # ----------------------------
  # Build n safely (non-vectorized column-existence checks)
  # ----------------------------
  if ("n_events" %in% names(annual_tbl_pred)) {
    annual_tbl_pred$n <- annual_tbl_pred$n_events
  } else if ("n_rows" %in% names(annual_tbl_pred)) {
    annual_tbl_pred$n <- annual_tbl_pred$n_rows
  } else {
    annual_tbl_pred$n <- NA_integer_
  }

  if ("n_events" %in% names(annual_tbl_obs)) {
    annual_tbl_obs$n <- annual_tbl_obs$n_events
  } else if ("n_rows" %in% names(annual_tbl_obs)) {
    annual_tbl_obs$n <- annual_tbl_obs$n_rows
  } else {
    annual_tbl_obs$n <- NA_integer_
  }

  # ---- modeled ----
  mod <- annual_tbl_pred %>%
    dplyr::transmute(
      Year      = as.integer(Year),
      analyte   = as.character(analyte),
      treatment = as.character(treatment),
      source    = "Modeled",

      load_mean,   load_low95,   load_high95,
      conc_mean,   conc_low95,   conc_high95,
      volume_mean, volume_low95, volume_high95,

      n = as.integer(n)
    )

  # ---- observed ----
  obs <- annual_tbl_obs %>%
    dplyr::transmute(
      Year      = as.integer(Year),
      analyte   = as.character(analyte),
      treatment = as.character(treatment),
      source    = "Observed",

      load_mean,   load_low95,   load_high95,
      conc_mean,   conc_low95,   conc_high95,
      volume_mean, volume_low95, volume_high95,

      n = as.integer(n)
    )

  dplyr::bind_rows(mod, obs) %>%
    dplyr::mutate(
      treatment = factor(treatment, levels = c("CT","MT","ST")),
      source    = factor(source, levels = c("Modeled","Observed"))
    ) %>%
    dplyr::arrange(analyte, Year, treatment, source)
}

# Combine modeled and observed annual tables
annual_both <- bind_observed_modeled(annual_tbl_pred, annual_tbl_obs)

```


```{r plot obs and mod}
plot_annual_by_treatment_obs_vs_mod <- function(
  annual_both,
  analyte_pick,
  metric = c("load","conc","volume"),
  dodge = 0.20
) {
  metric <- match.arg(metric)

  y_mean <- paste0(metric, "_mean")
  y_low  <- paste0(metric, "_low95")
  y_high <- paste0(metric, "_high95")

  stopifnot(all(c("Year","analyte","treatment","source", y_mean, y_low, y_high) %in% names(annual_both)))

  d <- annual_both %>%
    dplyr::filter(as.character(analyte) == analyte_pick) %>%
    dplyr::mutate(
      Year = as.integer(Year),
      treatment = factor(treatment, levels = c("CT","MT","ST")),
      source = factor(source, levels = c("Modeled","Observed"))
    ) %>%
    dplyr::arrange(Year, treatment, source)

  if (nrow(d) == 0) stop("No rows found for analyte = '", analyte_pick, "'.")

  ylab <- switch(
    metric,
    "load"   = "Annual load (g)",
    "conc"   = "Flow-weighted concentration (mg/L)",
    "volume" = "Annual volume (L)"
  )

  pd <- ggplot2::position_dodge(width = dodge)

  trt_cols <- c(
    CT = "#1f77b4",
    MT = "#ff7f0e",
    ST = "#2ca02c"
  )

  # Split for cleaner layering
  d_mod <- d %>% dplyr::filter(source == "Modeled")
  d_obs <- d %>% dplyr::filter(source == "Observed")

  ggplot2::ggplot() +
    # --- Modeled uncertainty (linerange) ---
    ggplot2::geom_linerange(
      data = d_mod,
      ggplot2::aes(
        x = Year,
        ymin = .data[[y_low]],
        ymax = .data[[y_high]],
        color = treatment
      ),
      position = pd,
      linewidth = 1.6,
      alpha = 0.45,
      show.legend = FALSE
    ) +

    # --- Modeled line ---
    ggplot2::geom_line(
      data = d_mod,
      ggplot2::aes(
        x = Year,
        y = .data[[y_mean]],
        color = treatment,
        group = treatment
      ),
      position = pd,
      linewidth = 1.2,
      alpha = 0.9
    ) +

    # --- Modeled points: filled with treatment color, black outline ---
    ggplot2::geom_point(
      data = d_mod,
      ggplot2::aes(
        x = Year,
        y = .data[[y_mean]],
        fill = treatment,
        shape = source              # NEW: adds "Modeled vs Observed" legend without changing appearance
      ),
      position = pd,
      #shape = 21,
      size = 3.4,
      stroke = 0.9,
      color = "black"
    ) +

    # --- Observed uncertainty (linerange) ---
    ggplot2::geom_linerange(
      data = d_obs,
      ggplot2::aes(
        x = Year,
        ymin = .data[[y_low]],
        ymax = .data[[y_high]],
        color = treatment
      ),
      position = pd,
      linewidth = 1.2,
      alpha = 0.35,
      show.legend = FALSE
    ) +

    # --- Observed points: white fill, treatment-colored outline ---
    ggplot2::geom_point(
      data = d_obs,
      ggplot2::aes(
        x = Year,
        y = .data[[y_mean]],
        color = treatment,
        shape = source              # NEW
      ),
      position = pd,
      #shape = 21,
      fill = "white",
      size = 2.8,
      stroke = 1.2
    ) +

    ggplot2::scale_color_manual(values = trt_cols, name = "Treatment") +
    ggplot2::scale_fill_manual(values = trt_cols, guide = "none") +

    # NEW: keep shapes identical in the plot, but create a legend for source
    ggplot2::scale_shape_manual(
      values = c(Modeled = 21, Observed = 21),
      name = "Estimate type",
      drop = FALSE
    ) +
    ggplot2::guides(
      shape = ggplot2::guide_legend(
        override.aes = list(
          fill  = c("black", "white"),
          color = c("black", "black"),
          stroke = c(0.9, 1.2),
          size = c(3.4, 2.8)
        )
      )
    ) +

    ggplot2::scale_x_continuous(
      breaks = sort(unique(d$Year)),
      expand = ggplot2::expansion(mult = c(0.01, 0.02))
    ) +
    ggplot2::labs(
      x = "Year",
      y = ylab,
      title = paste0(analyte_pick, ": annual ", metric, " by treatment")
    ) +
    ggplot2::theme_bw(base_size = 13) +
    ggplot2::theme(
      axis.text.x = ggplot2::element_text(angle = 45, hjust = 1),
      plot.title = ggplot2::element_text(face = "bold"),
      legend.position = "right",
      panel.grid.minor = ggplot2::element_blank()
    )
}

```

```{r execute plot both}
# Example usage:
plot_annual_by_treatment_obs_vs_mod(annual_both, analyte_pick = "NH4", metric = "load")
```

```{r save all plots}
# ------------------------------------------------------------
# Save images (v1p8)
# ------------------------------------------------------------
sanitize_slug <- function(x) {
  x <- as.character(x)
  x <- tolower(x)
  x <- gsub("[[:space:]]+", "_", x)
  x <- gsub("[^a-z0-9_]+", "", x)
  x <- gsub("_+", "_", x)
  x <- gsub("^_|_$", "", x)
  x
}

save_all_analyte_obs_vs_mod_plots <- function(
  annual_both,
  metric = c("load","conc","volume"),
  out_dir = file.path("../figs", "annual_obs_vs_modeled_v1p8"),
  width = 9,
  height = 5.5,
  dpi = 300,
  device = "jpg"
) {
  metric <- match.arg(metric)
  dir.create(out_dir, recursive = TRUE, showWarnings = FALSE)

  analytes <- sort(unique(as.character(annual_both$analyte)))

  log <- vector("list", length(analytes))

  for (i in seq_along(analytes)) {
    a <- analytes[i]
    slug <- sanitize_slug(a)

    p <- try(
      plot_annual_by_treatment_obs_vs_mod(
        annual_both = annual_both,
        analyte_pick = a,
        metric = metric
      ),
      silent = TRUE
    )

    if (inherits(p, "try-error")) {
      log[[i]] <- tibble::tibble(
        analyte = a,
        saved = FALSE,
        file = NA_character_,
        error = as.character(p)
      )
      next
    }

    fname <- file.path(out_dir, paste0("annual_", metric, "_", slug, "_obs_vs_modeled_v1p8.", device))

    ggplot2::ggsave(
      filename = fname,
      plot = p,
      width = width,
      height = height,
      dpi = dpi,
      device = device
    )

    log[[i]] <- tibble::tibble(
      analyte = a,
      saved = TRUE,
      file = fname,
      error = NA_character_
    )
  }

  dplyr::bind_rows(log)
}

# Example:
log_load <- save_all_analyte_obs_vs_mod_plots(annual_both, metric = "load")
# log_conc <- save_all_analyte_obs_vs_mod_plots(annual_both, metric = "conc")
# log_vol  <- save_all_analyte_obs_vs_mod_plots(annual_both, metric = "volume")

```

```{r annual-volume-obs-vs-modeled-plot}
# ============================================================
# Annual runoff volume by treatment: Observed vs Bayes Modeled
# Uses the same visual design as your load plots
# Converts to kiloliters (kL) for more manageable y-axis values
# ============================================================

stopifnot(exists("annual_combined_out") || exists("annual_both"))

# Prefer annual_combined_out from your export chunk (it is standardized)
vol_src <- if (exists("annual_combined_out")) annual_combined_out else annual_both

# Keep only the two sources we need (handle either naming convention)
vol_src <- vol_src %>%
  dplyr::mutate(
    source = dplyr::case_when(
      source %in% c("Bayes_Modeled", "Modeled") ~ "Modeled",
      source %in% c("Observed")                 ~ "Observed",
      TRUE ~ as.character(source)
    )
  ) %>%
  dplyr::filter(source %in% c("Modeled","Observed")) %>%
  dplyr::mutate(
    Year = as.integer(Year),
    treatment = factor(as.character(treatment), levels = c("CT","MT","ST")),
    source = factor(as.character(source), levels = c("Modeled","Observed"))
  )

# Volume is duplicated across analytes in your annual tables, so collapse safely.
# If it's truly identical across analytes, mean() will preserve it.
vol_annual <- vol_src %>%
  dplyr::group_by(Year, treatment, source) %>%
  dplyr::summarise(
    volume_mean   = mean(volume_mean,   na.rm = TRUE) / 1000,
    volume_low95  = mean(volume_low95,  na.rm = TRUE) / 1000,
    volume_high95 = mean(volume_high95, na.rm = TRUE) / 1000,
    .groups = "drop"
  ) %>%
  dplyr::arrange(Year, treatment, source) %>%
  dplyr::rename(
    volume_low  = volume_low95,
    volume_high = volume_high95
  )

if (nrow(vol_annual) == 0) stop("No annual volume rows found after filtering sources.")

# Split for cleaner layering (same as your load plot function)
d_mod <- vol_annual %>% dplyr::filter(source == "Modeled")
d_obs <- vol_annual %>% dplyr::filter(source == "Observed")

pd <- ggplot2::position_dodge(width = 0.20)

trt_cols <- c(
  CT = "#1f77b4",
  MT = "#ff7f0e",
  ST = "#2ca02c"
)

p_vol <- ggplot2::ggplot() +
  # Modeled uncertainty
  ggplot2::geom_linerange(
    data = d_mod,
    ggplot2::aes(x = Year, ymin = volume_low, ymax = volume_high, color = treatment),
    position = pd,
    linewidth = 1.6,
    alpha = 0.45,
    show.legend = FALSE
  ) +
  # Modeled line
  ggplot2::geom_line(
    data = d_mod,
    ggplot2::aes(x = Year, y = volume_mean, color = treatment, group = treatment),
    position = pd,
    linewidth = 1.2,
    alpha = 0.9
  ) +
  # Modeled points (filled, black outline) + legend mapping
  ggplot2::geom_point(
    data = d_mod,
    ggplot2::aes(x = Year, y = volume_mean, fill = treatment, shape = source),
    position = pd,
    size = 3.4,
    stroke = 0.9,
    color = "black"
  ) +
  # Observed uncertainty
  ggplot2::geom_linerange(
    data = d_obs,
    ggplot2::aes(x = Year, ymin = volume_low, ymax = volume_high, color = treatment),
    position = pd,
    linewidth = 1.2,
    alpha = 0.35,
    show.legend = FALSE
  ) +
  # Observed points (open, treatment outline) + legend mapping
  ggplot2::geom_point(
    data = d_obs,
    ggplot2::aes(x = Year, y = volume_mean, color = treatment, shape = source),
    position = pd,
    fill = "white",
    size = 2.8,
    stroke = 1.2
  ) +
  ggplot2::scale_color_manual(values = trt_cols, name = "Treatment") +
  ggplot2::scale_fill_manual(values = trt_cols, guide = "none") +
  ggplot2::scale_shape_manual(
    values = c(Modeled = 21, Observed = 21),
    name = "Estimate type",
    drop = FALSE
  ) +
  ggplot2::guides(
    shape = ggplot2::guide_legend(
      override.aes = list(
        fill   = c("black", "white"),
        color  = c("black", "black"),
        stroke = c(0.9, 1.2),
        size   = c(3.4, 2.8)
      )
    )
  ) +
  ggplot2::scale_x_continuous(
    breaks = sort(unique(vol_annual$Year)),
    expand = ggplot2::expansion(mult = c(0.01, 0.02))
  ) +
  ggplot2::labs(
    x = "Year",
    y = "Annual runoff volume (kL)",
    title = "Annual runoff volume by treatment"
  ) +
  ggplot2::theme_bw(base_size = 13) +
  ggplot2::theme(
    axis.text.x = ggplot2::element_text(angle = 45, hjust = 1),
    plot.title = ggplot2::element_text(face = "bold"),
    legend.position = "right",
    panel.grid.minor = ggplot2::element_blank()
  )

# ------------------------------------------------------------
# Save volume plot
# ------------------------------------------------------------

fig_dir <- file.path("../figs", paste0("annual_volume_", model_version))
dir.create(fig_dir, recursive = TRUE, showWarnings = FALSE)

fig_file <- file.path(
  "../figs",
  paste0("annual_volume_obs_vs_modeled_", model_version, ".jpg")
)

ggplot2::ggsave(
  filename = fig_file,
  plot = p_vol,
  width = 9,
  height = 5.5,
  dpi = 300
)

message("[OK] Wrote: ", fig_file)
p_vol
```

```{r export-annual-volume-wide-csv}
# ============================================================
# Export annual volume summary (Modeled + optional Observed)
# Row = Year
# Columns = CT/MT/ST with mean/low/high
# Also writes a publication-ready version: "mean (low, high)"
# Units: kL (converted from L by /1000)
# ============================================================

stopifnot(exists("annual_combined_out") || exists("annual_both"))
stopifnot(exists("model_version"))

vol_src <- if (exists("annual_combined_out")) annual_combined_out else annual_both

vol_src <- vol_src %>%
  dplyr::mutate(
    source = dplyr::case_when(
      source %in% c("Bayes_Modeled", "Modeled") ~ "Modeled",
      source %in% c("Observed")                 ~ "Observed",
      TRUE ~ as.character(source)
    )
  ) %>%
  dplyr::filter(source %in% c("Modeled","Observed")) %>%
  dplyr::mutate(
    Year = as.integer(Year),
    treatment = factor(as.character(treatment), levels = c("CT","MT","ST")),
    source = factor(as.character(source), levels = c("Modeled","Observed"))
  )

# Collapse across analytes (volume repeats per analyte)
vol_annual <- vol_src %>%
  dplyr::group_by(Year, treatment, source) %>%
  dplyr::summarise(
    volume_mean = mean(volume_mean, na.rm = TRUE) / 1000,
    volume_low  = mean(volume_low95,  na.rm = TRUE) / 1000,
    volume_high = mean(volume_high95, na.rm = TRUE) / 1000,
    .groups = "drop"
  )

out_dir <- "../out"
dir.create(out_dir, recursive = TRUE, showWarnings = FALSE)

# ---------- (A) Modeled wide export ----------
vol_modeled_wide <- vol_annual %>%
  dplyr::filter(source == "Modeled") %>%
  dplyr::mutate(col = paste0(as.character(treatment), "_mod")) %>%
  dplyr::select(Year, col, volume_mean, volume_low, volume_high) %>%
  tidyr::pivot_wider(
    names_from  = col,
    values_from = c(volume_mean, volume_low, volume_high),
    names_glue  = "{col}_{.value}"
  ) %>%
  dplyr::rename_with(~ gsub("_volume_", "_volume_kL_", .x, fixed = TRUE)) %>%
  dplyr::arrange(Year)

# ----- Add Total row (sum across years) -----
vol_modeled_wide <- vol_modeled_wide %>% dplyr::mutate(Year = as.character(Year))

vol_modeled_total <- vol_modeled_wide %>%
  dplyr::summarise(
    Year = "Total",
    dplyr::across(-Year, ~ sum(.x, na.rm = TRUE))
  )

vol_modeled_wide <- dplyr::bind_rows(vol_modeled_wide, vol_modeled_total)

f_modeled <- file.path(out_dir, paste0("annual_volume_kL_wide_modeled_", model_version, ".csv"))
readr::write_csv(vol_modeled_wide, f_modeled)
message("[OK] Wrote: ", f_modeled)

# ---------- (B) Pub-ready modeled export ----------
fmt_ci <- function(mean, low, high, digits = 0) {
  ifelse(
    is.finite(mean) & is.finite(low) & is.finite(high),
    paste0(
      formatC(mean, format = "f", digits = digits),
      " (",
      formatC(low,  format = "f", digits = digits),
      ", ",
      formatC(high, format = "f", digits = digits),
      ")"
    ),
    NA_character_
  )
}

vol_modeled_pub <- vol_modeled_wide %>%
  dplyr::transmute(
    Year,
    CT = fmt_ci(CT_mod_volume_kL_mean, CT_mod_volume_kL_low, CT_mod_volume_kL_high, digits = 0),
    MT = fmt_ci(MT_mod_volume_kL_mean, MT_mod_volume_kL_low, MT_mod_volume_kL_high, digits = 0),
    ST = fmt_ci(ST_mod_volume_kL_mean, ST_mod_volume_kL_low, ST_mod_volume_kL_high, digits = 0)
  )

# ----- Add Total row to pub table -----
vol_modeled_pub_total <- vol_modeled_wide %>%
  dplyr::filter(Year == "Total") %>%
  dplyr::transmute(
    Year = "Total",
    CT = fmt_ci(CT_mod_volume_kL_mean,
                CT_mod_volume_kL_low,
                CT_mod_volume_kL_high,
                digits = 0),
    MT = fmt_ci(MT_mod_volume_kL_mean,
                MT_mod_volume_kL_low,
                MT_mod_volume_kL_high,
                digits = 0),
    ST = fmt_ci(ST_mod_volume_kL_mean,
                ST_mod_volume_kL_low,
                ST_mod_volume_kL_high,
                digits = 0)
  )

vol_modeled_pub <- dplyr::bind_rows(
  vol_modeled_pub,
  vol_modeled_pub_total
)

f_modeled_pub <- file.path(out_dir, paste0("annual_volume_kL_wide_modeled_pub_", model_version, ".csv"))
readr::write_csv(vol_modeled_pub, f_modeled_pub)
message("[OK] Wrote: ", f_modeled_pub)

# ---------- Optional: Observed wide export too ----------
vol_observed_wide <- vol_annual %>%
  dplyr::filter(source == "Observed") %>%
  dplyr::mutate(col = paste0(as.character(treatment), "_obs")) %>%
  dplyr::select(Year, col, volume_mean, volume_low, volume_high) %>%
  tidyr::pivot_wider(
    names_from  = col,
    values_from = c(volume_mean, volume_low, volume_high),
    names_glue  = "{col}_{.value}"
  ) %>%
  dplyr::rename_with(~ gsub("_volume_", "_volume_kL_", .x, fixed = TRUE)) %>%
  dplyr::arrange(Year)

f_obs <- file.path(out_dir, paste0("annual_volume_kL_wide_observed_", model_version, ".csv"))
readr::write_csv(vol_observed_wide, f_obs)
message("[OK] Wrote: ", f_obs)

vol_observed_pub <- vol_observed_wide %>%
  dplyr::transmute(
    Year,
    CT = fmt_ci(CT_obs_volume_kL_mean, CT_obs_volume_kL_low, CT_obs_volume_kL_high, digits = 0),
    MT = fmt_ci(MT_obs_volume_kL_mean, MT_obs_volume_kL_low, MT_obs_volume_kL_high, digits = 0),
    ST = fmt_ci(ST_obs_volume_kL_mean, ST_obs_volume_kL_low, ST_obs_volume_kL_high, digits = 0)
  )

f_obs_pub <- file.path(out_dir, paste0("annual_volume_kL_wide_observed_pub_", model_version, ".csv"))
readr::write_csv(vol_observed_pub, f_obs_pub)
message("[OK] Wrote: ", f_obs_pub)

# Print the modeled pub-ready table in the Rmd
vol_modeled_pub

```


```{r export bayes annual predictions}
# ============================================================
# v1p8: Export annual posterior predictions (and optional observed)
# ============================================================

export_dir <- "../out"
dir.create(export_dir, recursive = TRUE, showWarnings = FALSE)

# ---- 1) Annual posterior summary (Modeled) ----
annual_pred_out <- annual_tbl_pred %>%
  dplyr::mutate(
    model_version = model_version,
    source = "Bayes_Modeled",
    interval_prob = dplyr::coalesce(prob_ci, 0.95)
  ) %>%
  dplyr::select(
    model_version, source, interval_prob,
    Year, analyte, treatment,
    n_events, n_rows,
    volume_mean, volume_low, volume_high,
    conc_mean,   conc_low,   conc_high,
    load_mean,   load_low,   load_high
  ) %>%
  dplyr::arrange(analyte, Year, treatment)

pred_file <- file.path(export_dir, paste0("annual_load_summary_bayes_", model_version, ".csv"))
readr::write_csv(annual_pred_out, pred_file)
message("[OK] Wrote: ", pred_file)

# ---- 2) Observed annual summary in same schema ----
pick_ci_cols <- function(df, prefix) {
  mean_nm  <- paste0(prefix, "_mean")
  low95_nm <- paste0(prefix, "_low95")
  high95_nm <- paste0(prefix, "_high95")
  low_nm   <- paste0(prefix, "_low")
  high_nm  <- paste0(prefix, "_high")

  if (all(c(mean_nm, low95_nm, high95_nm) %in% names(df))) {
    list(mean = mean_nm, low = low95_nm, high = high95_nm)
  } else if (all(c(mean_nm, low_nm, high_nm) %in% names(df))) {
    list(mean = mean_nm, low = low_nm, high = high_nm)
  } else {
    stop(
      sprintf(
        "Observed table missing CI columns for '%s'. Need (%s,%s,%s) or (%s,%s,%s). Present: %s",
        prefix,
        mean_nm, low95_nm, high95_nm,
        mean_nm, low_nm, high_nm,
        paste(names(df), collapse = ", ")
      )
    )
  }
}

vol_cols  <- pick_ci_cols(annual_tbl_obs, "volume")
conc_cols <- pick_ci_cols(annual_tbl_obs, "conc")
load_cols <- pick_ci_cols(annual_tbl_obs, "load")

annual_obs_out <- annual_tbl_obs %>%
  dplyr::mutate(
    model_version = model_version,
    source = "Observed",
    interval_prob = if ("prob_ci" %in% names(.)) dplyr::coalesce(prob_ci, 0.95) else 0.95
  ) %>%
  dplyr::transmute(
    model_version, source, interval_prob,
    Year,
    analyte   = as.character(analyte),
    treatment = as.character(treatment),
    n_events,
    n_rows = NA_integer_,

    volume_mean = .data[[vol_cols$mean]],
    volume_low  = .data[[vol_cols$low]],
    volume_high = .data[[vol_cols$high]],

    conc_mean = .data[[conc_cols$mean]],
    conc_low  = .data[[conc_cols$low]],
    conc_high = .data[[conc_cols$high]],

    load_mean = .data[[load_cols$mean]],
    load_low  = .data[[load_cols$low]],
    load_high = .data[[load_cols$high]]
  ) %>%
  dplyr::arrange(analyte, Year, treatment)

# ---- 3) Combined (Modeled + Observed) ----
annual_combined_out <- dplyr::bind_rows(annual_pred_out, annual_obs_out) %>%
  dplyr::arrange(source, analyte, Year, treatment)

comb_file <- file.path(export_dir, paste0("annual_load_summary_bayes_plus_observed_", model_version, ".csv"))
readr::write_csv(annual_combined_out, comb_file)
message("[OK] Wrote: ", comb_file)

# ---- 4) Optional: per-draw export (for Python fan charts / method-comparison) ----
export_annual_draws <- function(
  post,
  d_pred,
  years_used,
  analyte_lookup,
  block_lookup,
  sampler_lookup,
  flume_lookup,
  crop_lookup,
  c_stats,
  V_mean, V_sd,
  stir_mean, stir_sd,
  irr_mean, irr_sd,            # NEW v1p8
  n_draws = 400,
  seed = 1,
  drop_unmapped_analytes = TRUE,
  drop_missing_c_stats = TRUE,
  drop_unmapped_years = FALSE,
  drop_unmapped_other = FALSE
) {
  needed_cols <- c(
    "event_id","Year","Treatment","analyte_abbr","STIR_raw",
    "Rep","SampleMethod","FlumeMethod","irr_num","dup","cin_z",
    "Crop","RES_z"
  )
  miss <- setdiff(needed_cols, names(d_pred))
  if (length(miss) > 0) stop("d_pred is missing columns: ", paste(miss, collapse = ", "))

  set.seed(seed)

  # draw subsample
  n_total <- nrow(post$alpha)
  draw_idx <- if (n_total > n_draws) sample.int(n_total, n_draws) else seq_len(n_total)
  D <- length(draw_idx)

  # map indices + stats
  dat <- d_pred %>%
    dplyr::left_join(analyte_lookup %>% dplyr::select(A, analyte_abbr), by = "analyte_abbr") %>%
    dplyr::left_join(block_lookup   %>% dplyr::select(B, Rep),          by = "Rep") %>%
    dplyr::left_join(sampler_lookup %>% dplyr::select(S, SampleMethod), by = "SampleMethod") %>%
    dplyr::left_join(flume_lookup   %>% dplyr::select(Fu, FlumeMethod), by = "FlumeMethod") %>%
    dplyr::left_join(crop_lookup    %>% dplyr::select(Cr, Crop),        by = "Crop") %>%
    dplyr::left_join(c_stats %>% dplyr::select(A, c_mean, c_sd) %>% dplyr::distinct(), by = "A") %>%
    dplyr::mutate(
      Y      = match(Year, years_used),
      STIR_z = (STIR_raw - stir_mean) / stir_sd,
      irr_z  = (irr_num  - irr_mean)  / irr_sd     # NEW v1p8
    )

  # ---- Diagnose + optionally drop unmapped analytes ----
  if (anyNA(dat$A)) {
    bad_an <- sort(unique(dat$analyte_abbr[is.na(dat$A)]))
    msg <- paste0(
      "Unmapped analyte_abbr (not in analyte_lookup): ",
      paste(bad_an, collapse = ", "), "."
    )
    if (!isTRUE(drop_unmapped_analytes)) stop(msg)
    warning(msg, " Dropping these rows from d_pred for per-draw export.")
    dat <- dat %>% dplyr::filter(!is.na(A))
  }

  # ---- Diagnose + optionally drop missing c_stats ----
  if (anyNA(dat$c_mean) || anyNA(dat$c_sd)) {
    bad_an2 <- sort(unique(dat$analyte_abbr[is.na(dat$c_mean) | is.na(dat$c_sd)]))
    msg2 <- paste0(
      "Missing c_stats (c_mean/c_sd) for analytes: ",
      paste(bad_an2, collapse = ", "), "."
    )
    if (!isTRUE(drop_missing_c_stats)) stop(msg2)
    warning(msg2, " Dropping these rows from d_pred for per-draw export.")
    dat <- dat %>% dplyr::filter(!is.na(c_mean), !is.na(c_sd))
  }

  # ---- Years_used mismatch (optional drop) ----
  if (anyNA(dat$Y)) {
    bad_y <- sort(unique(dat$Year[is.na(dat$Y)]))
    msgy <- paste0("Some Year not found in years_used: ", paste(bad_y, collapse = ", "), ".")
    if (!isTRUE(drop_unmapped_years)) stop(msgy)
    warning(msgy, " Dropping these rows from d_pred for per-draw export.")
    dat <- dat %>% dplyr::filter(!is.na(Y))
  }

  # ---- Other lookups (optional drop; default: strict) ----
  if (anyNA(dat$B) || anyNA(dat$S) || anyNA(dat$Fu) || anyNA(dat$Cr)) {
    parts <- c()
    if (anyNA(dat$B))  parts <- c(parts, paste0("Rep: ", paste(sort(unique(dat$Rep[is.na(dat$B)])), collapse = ", ")))
    if (anyNA(dat$S))  parts <- c(parts, paste0("SampleMethod: ", paste(sort(unique(dat$SampleMethod[is.na(dat$S)])), collapse = ", ")))
    if (anyNA(dat$Fu)) parts <- c(parts, paste0("FlumeMethod: ", paste(sort(unique(dat$FlumeMethod[is.na(dat$Fu)])), collapse = ", ")))
    if (anyNA(dat$Cr)) parts <- c(parts, paste0("Crop: ", paste(sort(unique(dat$Crop[is.na(dat$Cr)])), collapse = ", ")))
    msgo <- paste0("Index mapping produced NAs for: ", paste(parts, collapse = " | "), ".")

    if (!isTRUE(drop_unmapped_other)) stop(msgo)
    warning(msgo, " Dropping these rows from d_pred for per-draw export.")
    dat <- dat %>% dplyr::filter(!is.na(B), !is.na(S), !is.na(Fu), !is.na(Cr))
  }

  if (anyNA(dat$irr_z)) stop("Some irr_z are NA. Check irr_mean/irr_sd and irr_num mapping.")

  # Integer indices
  dat <- dat %>%
    dplyr::mutate(
      A  = as.integer(A),
      B  = as.integer(B),
      S  = as.integer(S),
      Fu = as.integer(Fu),
      Y  = as.integer(Y),
      Cr = as.integer(Cr)
    )

  groups <- dat %>%
    dplyr::distinct(Year, analyte_abbr, Treatment) %>%
    dplyr::arrange(analyte_abbr, Year, Treatment)

  out <- vector("list", length = D * nrow(groups))
  kk <- 1L

  for (dd in seq_len(D)) {
    s <- draw_idx[dd]

    for (g in seq_len(nrow(groups))) {

      yr <- groups$Year[g]
      an <- groups$analyte_abbr[g]
      tr <- groups$Treatment[g]

      dsub <- dat %>% dplyr::filter(Year == yr, analyte_abbr == an, Treatment == tr)
      if (nrow(dsub) == 0) next

      # de-duplicate to one row per event_id for analyte
      dsub1 <- dsub %>% dplyr::distinct(event_id, .keep_all = TRUE)

      A  <- dsub1$A;  B <- dsub1$B;  Sidx <- dsub1$S;  Fu <- dsub1$Fu;  Y <- dsub1$Y;  Cr <- dsub1$Cr
      stz <- dsub1$STIR_z
      cin <- dsub1$cin_z
      irz <- dsub1$irr_z               # NEW v1p8
      dup <- dsub1$dup
      res <- dsub1$RES_z
      c_mean <- dsub1$c_mean
      c_sd   <- dsub1$c_sd

      ev <- dsub %>%
        dplyr::distinct(event_id, STIR_z, RES_z, Crop) %>%
        dplyr::left_join(crop_lookup %>% dplyr::select(Cr, Crop), by = "Crop") %>%
        dplyr::mutate(Cr = as.integer(Cr)) %>%
        dplyr::filter(!is.na(Cr))

      muV_z_ev <- post$a_V[s] +
        post$b_V[s]         * ev$STIR_z +
        post$beta_res_V[s]  * ev$RES_z +
        post$gamma_Cr_V[s, ev$Cr]

      Vsum <- sum(V_mean + V_sd * muV_z_ev, na.rm = TRUE)

      muV_z_row <- post$a_V[s] +
        post$b_V[s]         * stz +
        post$beta_res_V[s]  * res +
        post$gamma_Cr_V[s, Cr]

      gB <- post$gamma_B[s, , , drop = TRUE][cbind(A, B)]
      gS <- post$gamma_S[s, , , drop = TRUE][cbind(A, Sidx)]
      gF <- post$gamma_F[s, , , drop = TRUE][cbind(A, Fu)]
      kY <- post$F_year[s, , , drop = TRUE][cbind(Y, A)]
      gCrC <- post$gamma_Cr[s, , , drop = TRUE][cbind(A, Cr)]

      muC_z <- post$alpha[s, A] +
        post$beta_stir[s, A]  * stz +
        post$beta_cin[s, A]   * cin +
        post$beta_vol[s]      * muV_z_row +
        post$beta_irr[s, A]   * irz +      # NEW v1p8
        post$beta_dup[s, A]   * dup +
        post$beta_res_C[s, A] * res +
        gCrC + gB + gS + gF + kY

      C_raw <- c_mean + c_sd * muC_z
      V_row_raw <- V_mean + V_sd * muV_z_row

      Lmg <- sum(C_raw * V_row_raw, na.rm = TRUE)
      Lg  <- Lmg / 1000
      Cfw <- if (is.finite(Vsum) && Vsum > 0) ((Lg * 1000) / Vsum) else NA_real_

      out[[kk]] <- tibble::tibble(
        draw = dd,
        draw_id = s,
        Year = yr,
        analyte = an,
        treatment = tr,
        volume_L = Vsum,
        conc_fw_mg_L = Cfw,
        load_g = Lg
      )
      kk <- kk + 1L
    }
  }

  dplyr::bind_rows(out[seq_len(kk - 1L)]) %>%
    dplyr::arrange(analyte, Year, treatment, draw)
}
```

```{r study-period-load-table-kg}
# ============================================================
# Study-period total loads by treatment (Modeled) in kg
# ============================================================

stopifnot(exists("annual_pred_out"))
stopifnot(all(c("analyte","treatment","load_mean","load_low","load_high") %in% names(annual_pred_out)))

study_load_tbl <- annual_pred_out %>%
  dplyr::filter(source == "Bayes_Modeled") %>%
  dplyr::group_by(analyte, treatment) %>%
  dplyr::summarise(
    load_sum_mean_kg = sum(load_mean, na.rm = TRUE) / 1000,
    load_sum_low_kg  = sum(load_low,  na.rm = TRUE) / 1000,
    load_sum_high_kg = sum(load_high, na.rm = TRUE) / 1000,
    .groups = "drop"
  ) %>%
  dplyr::mutate(
    treatment = factor(as.character(treatment), levels = c("CT","MT","ST"))
  ) %>%
  dplyr::arrange(analyte, treatment)

study_load_wide <- study_load_tbl %>%
  dplyr::mutate(col = paste0(as.character(treatment), "_mod")) %>%
  dplyr::select(analyte, col,
                load_sum_mean_kg,
                load_sum_low_kg,
                load_sum_high_kg) %>%
  tidyr::pivot_wider(
    names_from  = col,
    values_from = c(load_sum_mean_kg,
                    load_sum_low_kg,
                    load_sum_high_kg),
    names_glue  = "{col}_{.value}"
  ) %>%
  dplyr::arrange(analyte)

# ---- Save to /out ----
out_file <- file.path("../out",
                      paste0("study_period_total_loads_kg_", model_version, ".csv"))

readr::write_csv(study_load_wide, out_file)
message("[OK] Wrote: ", out_file)

study_load_wide

# ============================================================
# Publication-ready summary (one column per treatment)
# Cells: "mean (low, high)" in kg
# ============================================================

fmt_ci <- function(mean, low, high, digits = 2) {
  ifelse(
    is.finite(mean) & is.finite(low) & is.finite(high),
    paste0(
      formatC(mean, format = "f", digits = digits),
      " (",
      formatC(low,  format = "f", digits = digits),
      ", ",
      formatC(high, format = "f", digits = digits),
      ")"
    ),
    NA_character_
  )
}

study_load_pub <- study_load_wide %>%
  dplyr::transmute(
    analyte,
    CT = fmt_ci(CT_mod_load_sum_mean_kg, CT_mod_load_sum_low_kg, CT_mod_load_sum_high_kg, digits = 2),
    MT = fmt_ci(MT_mod_load_sum_mean_kg, MT_mod_load_sum_low_kg, MT_mod_load_sum_high_kg, digits = 2),
    ST = fmt_ci(ST_mod_load_sum_mean_kg, ST_mod_load_sum_low_kg, ST_mod_load_sum_high_kg, digits = 2)
  )

pub_file <- file.path("../out", paste0("study_period_total_loads_kg_pub_", model_version, ".csv"))
readr::write_csv(study_load_pub, pub_file)
message("[OK] Wrote: ", pub_file)

study_load_pub
```


```{r export bayes annual predictions}
# Uncomment to write per-draw table (can be big):
annual_draws <- export_annual_draws(
  post = post,
  d_pred = d_pred,
  years_used = years_used,
  analyte_lookup = analyte_lookup,
  block_lookup = block_lookup,
  sampler_lookup = sampler_lookup,
  flume_lookup = flume_lookup,
  crop_lookup = crop_lookup,
  c_stats = c_stats,
  V_mean = V_mean, V_sd = V_sd,
  stir_mean = stir_mean, stir_sd = stir_sd,
  irr_mean = irr_mean, irr_sd = irr_sd,
  n_draws = 400,
  seed = 1
)

# Save to repo-level out folder (one level up from /code)
export_dir <- "../out"
dir.create(export_dir, showWarnings = FALSE, recursive = TRUE)

draws_file <- file.path(
  export_dir,
  paste0("annual_load_draws_bayes_", model_version, ".csv")
)

readr::write_csv(annual_draws, draws_file)
message("[OK] Wrote: ", draws_file)

```
